---
title: "GLMM Aerial Effort Estimation"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{GLMM Aerial Effort Estimation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## When to Use the GLMM Estimator

If your pilot always flies at the same time of day — say, 10 AM — your
instantaneous count systematically over- or under-represents total daily effort.
Early-morning flights catch fewer anglers than are present at peak hours;
late-afternoon flights may miss early starters entirely. When flight timing is
non-random, the simple expansion (count × h_open / v) inherits this temporal
bias.

The GLMM approach (Askey et al. 2018) models how angler counts change through
the day using a quadratic hour effect and a day-level random intercept:
`count ~ poly(hour, 2) + (1 | date)`. Once the diurnal curve is estimated, the
model integrates predicted counts across the full open-water window — correcting
for wherever in the curve the actual flight fell.

**When to use the GLMM estimator:**

1. Flights always occur in the same part of the day (fixed morning or afternoon
   schedule).
2. You have multiple overflights per day across several days (minimum 8–10
   survey days recommended for stable random-effect estimation).

**When to use the simple estimator:**

1. Flight timing is randomly assigned across the open-water window.
2. You have only one count per day.

For the basic aerial workflow without GLMM correction, see the
[aerial surveys vignette](aerial-surveys.html).

## Example Data

The `example_aerial_glmm_counts` dataset contains 12 survey days with 4
overflights per day at fixed hours (7, 10, 13, and 16 hours), producing
48 observations. Angler counts follow a diurnal curve with day-level Poisson
variability — representative of a scenario where a fixed morning-to-afternoon
flight schedule is used throughout the season.

```{r data}
library(tidycreel)
data(example_aerial_glmm_counts)
head(example_aerial_glmm_counts)
```

The four columns are `date`, `day_type`, `n_anglers` (instantaneous count), and
`time_of_flight` (decimal hour of each overflight).

## Building the Aerial Design

Build an aerial `creel_design` from the survey dates and attach the count data.
The `h_open = 14` argument specifies the number of hours the fishery is open
each day — this enters the final expansion after the diurnal curve is
integrated.

```{r design}
aerial_cal <- unique(example_aerial_glmm_counts[, c("date", "day_type")])
aerial_cal <- aerial_cal[order(aerial_cal$date), ]

design <- creel_design(
  aerial_cal,
  date        = date,
  strata      = day_type,
  survey_type = "aerial",
  h_open      = 14
)

design <- add_counts(design, example_aerial_glmm_counts)
print(design)
```

## GLMM Effort Estimation

Call `estimate_effort_aerial_glmm()` with `time_col = time_of_flight`. The
default model fits a negative-binomial GLMM with a quadratic temporal effect
and a day-level random intercept:

```{r glmm-default}
glmm_result <- estimate_effort_aerial_glmm(design, time_col = time_of_flight)
print(glmm_result)
```

The model fits the diurnal count curve over all 48 observations, then
numerically integrates the predicted mean count across the full open-water
window (from 0.5 hours before the earliest flight to 0.5 hours after the
latest). The integrated area is scaled by `h_open / visibility_correction` to
convert counts to angler-hours.

## Variance and Confidence Intervals

Two variance methods are available.

**Delta method (default):** Propagates the fixed-effect covariance matrix from
`lme4::vcov()` to the derived integral via a gradient vector. This is fast
and analytic, and is the default when `boot = FALSE`.

**Parametric bootstrap:** `lme4::bootMer()` re-fits the model under parametric
resampling `nsim` times and uses the SD of the resulting totals as the SE.
This method can give more accurate CIs for skewed count distributions. Use it
for final production analyses; the delta method is appropriate for exploratory
work.

```{r glmm-boot, eval = FALSE}
# Bootstrap CIs — use nboot = 500 for production analyses
glmm_boot <- estimate_effort_aerial_glmm(
  design,
  time_col = time_of_flight,
  boot = TRUE,
  nboot = 100L
)
print(glmm_boot)
```

## Downstream Estimation

GLMM effort feeds directly into the standard downstream estimators.
Attach interview data and call `estimate_catch_rate()` and
`estimate_total_catch()` on the same design object.

```{r downstream}
# Build a complementary design with matching interview dates
aerial_int_cal <- unique(example_aerial_counts[, c("date", "day_type")])
aerial_int_cal <- aerial_int_cal[order(aerial_int_cal$date), ]

design_int <- creel_design(
  aerial_int_cal,
  date        = date,
  strata      = day_type,
  survey_type = "aerial",
  h_open      = 14
)
design_int <- add_counts(design_int, example_aerial_counts)
design_int <- add_interviews(design_int, example_aerial_interviews,
  catch       = walleye_catch,
  effort      = hours_fished,
  trip_status = trip_status
)
catch_rate <- estimate_catch_rate(design_int)
print(catch_rate)

total_catch <- estimate_total_catch(design_int)
print(total_catch)
```

## Comparison: Simple vs. GLMM Estimator

Both estimators are applied to the same design. The simple estimator treats
each overflight count as a representative sample of the full open-water window
and expands by `h_open / v` directly. The GLMM estimator corrects for the
fixed-hour flight schedule by integrating the fitted diurnal curve.

```{r comparison}
# Simple aerial estimator — no diurnal correction
simple_result <- estimate_effort(design)

# GLMM result from above
# glmm_result already computed

# Side-by-side comparison
comparison <- rbind(
  data.frame(
    method = "GLMM",
    estimate = glmm_result$estimates$estimate,
    se = glmm_result$estimates$se,
    ci_lower = glmm_result$estimates$ci_lower,
    ci_upper = glmm_result$estimates$ci_upper,
    stringsAsFactors = FALSE
  ),
  data.frame(
    method = "Simple",
    estimate = simple_result$estimates$estimate,
    se = simple_result$estimates$se,
    ci_lower = simple_result$estimates$ci_lower,
    ci_upper = simple_result$estimates$ci_upper,
    stringsAsFactors = FALSE
  )
)

print(comparison)
```

The GLMM estimate corrects for the fact that all flights occurred at fixed hours
(7, 10, 13, 16). The simple estimator treats each count as representative of
the full open-water window, which inflates or deflates the total depending on
where the peak falls in the diurnal curve.

## Custom Formula

For surveys where a linear temporal term is sufficient — or where the analyst
prefers to control the model structure directly — pass a custom `formula`. The
formula must reference the actual count column (`n_anglers`) and the time
column by its exact name in the data.

```{r custom-formula}
glmm_linear <- estimate_effort_aerial_glmm(
  design,
  time_col = time_of_flight,
  formula  = n_anglers ~ time_of_flight + (1 | date)
)
print(glmm_linear)
```

A linear temporal term reduces flexibility but can improve stability when only
a few survey days are available. Use the default quadratic formula when you
have 8 or more survey days.

## References

- Askey, P. J., Parkinson, E. A., Rehill, P., & Post, J. R. (2018).
  Correcting for non-random flight timing in aerial creel surveys using a
  generalized linear mixed model. *North American Journal of Fisheries
  Management*, 38(5), 1204–1215.
  [https://doi.org/10.1002/nafm.10010](https://doi.org/10.1002/nafm.10010)

- Jones, C. M., & Pollock, K. H. (2012). Recreational survey methods:
  estimation of effort, harvest, and abundance. Chapter 19 in
  *Fisheries Techniques* (3rd ed.), pp. 883–919. American Fisheries Society.
