---
title: "Bus-Route Surveys"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Bus-Route Surveys}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## What is a Bus-Route Survey?

A bus-route survey visits multiple access points (sites) in a fixed route, called a
circuit. Unlike an access-point creel survey that stations an interviewer at one location
for the day, a bus-route crew travels the circuit repeatedly, counting and interviewing
anglers at each site as they pass through.

Sites are not sampled uniformly. Instead, each site is assigned a selection probability
proportional to its expected fishing effort: high-use sites appear in the sampling frame
with higher probability and are therefore visited more often, while low-use sites appear
with lower probability. This deliberate, probability-proportional-to-size design is what
makes valid inference possible.

Two types of data are collected during each visit:

- **Enumeration counts** — every angler party at the site is counted (`n_counted`).
- **Interviews** — a subset of parties is interviewed to collect effort and catch data
  (`n_interviewed`).

The ratio `n_counted / n_interviewed` (the enumeration expansion factor) scales the
interviewed catch and effort up to represent all parties at the site, including those not
interviewed. Without this expansion, busy sites are systematically underrepresented in
the total estimate.

The statistical framework follows Malvestuto et al. (1978) and Jones & Pollock (2012,
Chapter 19). tidycreel implements the general Horvitz-Thompson estimators from that
chapter directly.

## Key Concepts

### Inclusion Probability (πᵢ)

The inclusion probability πᵢ is the probability that a given sampling unit (site × period
combination) is included in the sample. For a two-stage bus-route design:

$$\pi_i = p\_site_i \times p\_period$$

where `p_site` is the probability of selecting site _i_ in a given pass, and `p_period`
is the probability that the period containing the visit is sampled. Sites with higher
expected effort get higher `p_site` values, meaning they are selected more often — and
correspondingly receive a lower inverse-probability weight in the estimator.

**Critical point:** πᵢ is a property of the *sampling design*, not of the site or the
anglers. It is specified when you design the survey, not computed from the interview data.

### Enumeration Counts

During each circuit pass, the crew counts every visible angler party at the site
(`n_counted`) and then interviews as many as possible (`n_interviewed`). When not all
parties are interviewed, the expansion factor

$$\text{expansion} = \frac{n\_counted}{n\_interviewed}$$

rescales the interviewed data so it represents all parties. If all parties are interviewed
(n_counted = n_interviewed), the expansion is 1 and no adjustment is needed. This is the
case in Malvestuto (1996) Box 20.6 Example 1, used in this vignette.

### Horvitz-Thompson Estimator

The total effort estimator (Jones & Pollock 2012, Eq. 19.4) is:

$$\hat{E} = \sum_i \frac{e_i}{\pi_i}$$

where $e_i = \text{hours\_fished}_i \times \text{expansion}_i$ is the enumeration-expanded
effort for interview _i_. Each site's contribution is weighted by the reciprocal of its
inclusion probability. A site sampled with probability 0.20 receives weight 5 — five times
the weight of a site sampled with probability 1.0. This inverse-probability weighting is
what makes the estimator statistically unbiased.

The harvest estimator (Jones & Pollock 2012, Eq. 19.5) follows the same structure:

$$\hat{H} = \sum_i \frac{h_i}{\pi_i}$$

## A Note on Existing Implementations

Some R packages that support bus-route surveys hardcode $\pi_i = 0.5$ or compute the
inclusion probability as `wait_time / circuit_time`. Neither is the inclusion probability
as defined by Horvitz-Thompson theory or by Jones & Pollock (2012). Using an incorrect
$\pi_i$ produces biased estimates whose magnitude of bias grows with the heterogeneity of
site probabilities.

tidycreel computes $\pi_i = p\_site \times p\_period$ directly from the sampling frame
you supply to `creel_design()`. This matches the primary source definition in Jones &
Pollock (2012, p. 912) and the worked example in Malvestuto (1996, Box 20.6). The
implementation is validated against that published example in the package test suite.

## Step-by-Step Example

We demonstrate the full bus-route workflow using the data from Malvestuto (1996) Box 20.6,
Example 1 — a four-site, single-circuit survey with 15 interviews over four days.

### Step 1: Define the Sampling Design

The sampling frame specifies the four sites, their selection probabilities (`p_site`), and
the period sampling probability (`p_period`). Site probabilities must sum to 1.0 within
each circuit.

```{r design}
library(tidycreel)

# Sampling frame: 4 sites, 1 circuit
# p_site values are proportional to expected fishing effort at each site
sampling_frame <- data.frame(
  site     = c("A", "B", "C", "D"),
  circuit  = "circ1",
  p_site   = c(0.30, 0.25, 0.40, 0.05), # must sum to 1.0
  p_period = 0.50 # uniform period sampling
)
# Inclusion probability: pi_i = p_site * p_period
# Site A: 0.30 * 0.50 = 0.15
# Site B: 0.25 * 0.50 = 0.125
# Site C: 0.40 * 0.50 = 0.20
# Site D: 0.05 * 0.50 = 0.025

# Survey calendar: 4 weekdays
calendar <- data.frame(
  date     = as.Date(c("2024-06-01", "2024-06-02", "2024-06-03", "2024-06-04")),
  day_type = "weekday"
)

# Build the bus-route design
design <- creel_design(
  calendar,
  date = date,
  strata = day_type,
  survey_type = "bus_route",
  sampling_frame = sampling_frame,
  site = site,
  circuit = circuit,
  p_site = p_site,
  p_period = p_period
)
print(design)
```

The design object stores the sampling frame and precomputes πᵢ for each site×circuit
combination. Site C has the highest selection probability (p_site = 0.40) and therefore
the highest πᵢ (0.20), which means its interviews will carry lower inverse-probability
weight than Site D's (πᵢ = 0.025).

### Step 2: Attach Interview Data

Each row of interview data represents one interviewed party. The `n_counted` and
`n_interviewed` columns record how many parties were at the site vs. how many were
actually interviewed. In this example both are equal (no expansion needed), but the
columns must still be supplied for bus-route designs.

```{r interviews}
# 15 interviews across 4 sites and 4 days
# Site A: 4 interviews (all 4 parties interviewed)
# Site B: 3 interviews (all 3 parties)
# Site C: 6 interviews (all 6 parties) — highest effort site
# Site D: 2 interviews (all 2 parties)
interviews_raw <- data.frame(
  date = as.Date(c(
    "2024-06-01", "2024-06-01", "2024-06-02", "2024-06-02", # Site A
    "2024-06-01", "2024-06-02", "2024-06-03", # Site B
    "2024-06-01", "2024-06-02", "2024-06-03", # Site C (rows 1-3)
    "2024-06-03", "2024-06-04", # Site C (rows 4-5)
    "2024-06-03", "2024-06-04", # Site D
    "2024-06-03" # Site C (row 6)
  )),
  day_type = "weekday",
  site = c(
    "A", "A", "A", "A",
    "B", "B", "B",
    "C", "C", "C",
    "C", "C",
    "D", "D",
    "C"
  ),
  circuit = "circ1",
  hours_fished = c(
    7.5, 7.5, 7.5, 7.5, # Site A: 4 x 7.5 = 30 h total
    20.0 / 3, 20.0 / 3, 20.0 / 3, # Site B: 3 x 6.67 = 20 h total
    57.5 / 6, 57.5 / 6, 57.5 / 6, # Site C rows 1-3
    57.5 / 6, 57.5 / 6, # Site C rows 4-5
    2.5, 2.5, # Site D: 2 x 2.5 = 5 h total
    57.5 / 6 # Site C row 6: total = 57.5 h
  ),
  fish_kept = c(
    1L, 0L, 1L, 0L,
    1L, 0L, 1L,
    1L, 1L, 0L,
    1L, 0L,
    0L, 0L,
    1L
  ),
  fish_caught = c(
    2L, 1L, 1L, 1L,
    2L, 1L, 1L,
    2L, 1L, 1L,
    1L, 1L,
    1L, 1L,
    1L
  ),
  n_counted = c(
    4L, 4L, 4L, 4L,
    3L, 3L, 3L,
    6L, 6L, 6L,
    6L, 6L,
    2L, 2L,
    6L
  ),
  n_interviewed = c(
    4L, 4L, 4L, 4L,
    3L, 3L, 3L,
    6L, 6L, 6L,
    6L, 6L,
    2L, 2L,
    6L
  ),
  trip_status = "complete"
)

# Attach interviews — pi_i is joined automatically from the sampling frame
design <- add_interviews(
  design,
  interviews_raw,
  effort        = hours_fished,
  catch         = fish_caught,
  harvest       = fish_kept,
  n_counted     = n_counted,
  n_interviewed = n_interviewed,
  trip_status   = trip_status
)
print(design)
```

`add_interviews()` joins πᵢ from the sampling frame to each interview row (stored as
`.pi_i`) and computes the enumeration expansion factor (`.expansion = n_counted /
n_interviewed`). These are used internally by the estimation functions.

### Step 3: Estimate Effort

`estimate_effort()` dispatches to the bus-route estimator and computes
$\hat{E} = \sum(e_i / \pi_i)$.

```{r effort}
# Estimate total angler-hours
effort_est <- estimate_effort(design)
print(effort_est)

# Inspect per-site contributions to the total
site_contribs <- get_site_contributions(effort_est)
print(site_contribs)
```

The site contributions table shows exactly how each site's effort feeds into the total.
Site C (πᵢ = 0.20, total effort = 57.5 h) contributes 57.5 / 0.20 = 287.5 angler-hours
to the total — the largest single-site contribution, reflecting its high effort despite
the relatively frequent sampling.

Site D (πᵢ = 0.025, total effort = 5 h) contributes 5 / 0.025 = 200 angler-hours. Though
Site D has little observed effort, its low sampling probability means it represents a much
larger population of fishing activity.

The total E_hat = 847.5 angler-hours matches Malvestuto (1996) Box 20.6 Example 1
exactly.

### Step 4: Estimate Harvest

`estimate_harvest_rate()` applies the same Horvitz-Thompson logic to fish kept.

```{r harvest}
# Estimate total fish kept (harvest)
harvest_est <- estimate_harvest_rate(design)
print(harvest_est)

# Estimate total catch (kept + released)
catch_est <- estimate_total_catch(design)
print(catch_est)
```

## Validation

The effort estimate above (E_hat = 847.5 angler-hours) reproduces the published result
from Malvestuto (1996) Box 20.6 Example 1 exactly. This agreement is verified by the
package test suite in `tests/testthat/test-primary-source-validation.R`. Each of the four
site contributions matches:

| Site | e_i (h) | πᵢ    | e_i / πᵢ |
|------|---------|-------|----------|
| A    | 30.0    | 0.150 | 200.0    |
| B    | 20.0    | 0.125 | 160.0    |
| C    | 57.5    | 0.200 | 287.5    |
| D    |  5.0    | 0.025 | 200.0    |
| **Total** | | | **847.5** |

## References

- Jones, C. M., & Pollock, K. H. (2012). Recreational survey methods: estimation of effort,
  harvest, and abundance. Chapter 19 in *Fisheries Techniques* (3rd ed.), pp. 883–919.
  American Fisheries Society.

- Malvestuto, S. P. (1996). Sampling the recreational angler. Chapter 20 in *Fisheries
  Techniques* (2nd ed.), pp. 591–623. American Fisheries Society.

- Malvestuto, S. P., Davies, W. D., & Shelton, W. L. (1978). An evaluation of the roving
  creel survey with nonuniform probability sampling. *Transactions of the American Fisheries
  Society*, 107(2), 255–262.
