---
title: "Camera Survey Analysis with tidycreel"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Camera Survey Analysis with tidycreel}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Overview

Remote cameras mounted at boat launches or trail heads record angler arrivals
with no observer present. This enables 24-hour coverage of access points that
would otherwise require continuous staffing. The `tidycreel` package supports
two camera-based sub-modes:

- **Counter mode** — the camera records a single daily total of incoming
  anglers (e.g., a passive infrared counter). The data arrive as one row per
  sampling day with a numeric ingress count.
- **Ingress-egress mode** — the camera records individual arrival and
  departure timestamps for each angler or party. The raw data are paired
  POSIXct timestamps that are preprocessed with
  `preprocess_camera_timestamps()` before entering the standard effort
  estimation workflow.

Both modes share a key concept: the `camera_status` column. Unlike random
missed observations (modeled via `missing_sections`), a camera failure is an
**informative gap** — the number of anglers that passed during the outage is
unknown and cannot be estimated from the surrounding data. These rows must be
removed before calling `add_counts()`.

## Example Data

This vignette uses three built-in datasets representing a hypothetical summer
creel survey at a Nebraska reservoir boat launch in June 2024.

```{r data}
library(tidycreel)

data(example_camera_counts)
data(example_camera_timestamps)
data(example_camera_interviews)

head(example_camera_counts)
head(example_camera_timestamps)
head(example_camera_interviews)
```

The `example_camera_counts` dataset contains 10 rows: nine operational days
plus one battery failure gap. The `example_camera_timestamps` dataset has 14
raw arrival/departure pairs across four sampling days. The
`example_camera_interviews` dataset has 40 complete angler interviews
targeting walleye and bass across eight sampling days.

## Counter Mode

### Build a Survey Calendar

Counter-mode surveys require a calendar that covers the sampling frame. Here
we build one from the unique dates in the counts and interview data.

```{r counter-calendar}
# Collect all unique sampling dates across datasets
all_dates <- sort(unique(c(
  example_camera_counts$date,
  example_camera_interviews$date
)))

# Assign day type for each date (weekday = Mon-Fri, weekend = Sat-Sun)
cam_calendar <- data.frame(
  date = all_dates,
  day_type = ifelse(
    weekdays(all_dates) %in% c("Saturday", "Sunday"),
    "weekend", "weekday"
  ),
  stringsAsFactors = FALSE
)

head(cam_calendar)
```

### Design Construction

Build a camera design using `creel_design()` with `survey_type = "camera"`.
The `camera_mode` argument is required. Omitting it produces an informative
error:

```{r counter-error, error = TRUE}
creel_design(
  cam_calendar,
  date = date, strata = day_type,
  survey_type = "camera"
)
```

Build the correct design with `camera_mode = "counter"`:

```{r counter-design}
design_counter <- creel_design(
  cam_calendar,
  date        = date,
  strata      = day_type,
  survey_type = "camera",
  camera_mode = "counter"
)
print(design_counter)
```

## Handling Informative Gaps

The battery failure row on 2024-06-11 has `ingress_count = NA`. This is not
a random unsampled day — the camera was physically unable to record. Including
it in `add_counts()` would silently propagate a missing value into the
Horvitz-Thompson estimator.

```{r gap-row}
# The gap row
subset(example_camera_counts, camera_status != "operational")
```

The correct approach is to **filter to operational rows** before calling
`add_counts()`. This is fundamentally different from `missing_sections`, which
models probabilistic non-coverage within a sampled period. A camera failure
means no data exist — the effort during that period is unknown.

```{r counter-filter}
# Keep only days when the camera was working
counts_clean <- subset(example_camera_counts, camera_status == "operational")
nrow(counts_clean) # 9 operational rows
```

### Effort Estimation

```{r counter-effort}
design_counter <- add_counts(design_counter, counts_clean)
effort_counter <- suppressWarnings(estimate_effort(design_counter))
print(effort_counter)
```

The `estimate` column is the Horvitz-Thompson estimate of total ingress angler
visits across the full survey period, with a 95% confidence interval.

## Ingress-Egress Mode

When the camera records individual arrival and departure timestamps, use
`preprocess_camera_timestamps()` to aggregate the raw pairs into daily effort
hours before calling `add_counts()`.

### Preprocess Timestamps

```{r ie-preprocess}
daily_effort <- preprocess_camera_timestamps(
  example_camera_timestamps,
  date_col    = date,
  ingress_col = ingress_time,
  egress_col  = egress_time
)
print(daily_effort)
```

`preprocess_camera_timestamps()` sums all valid trip durations within each day
and returns a data frame with `date` and `daily_effort_hours`. A warning is
issued for any rows where `egress_time < ingress_time` (negative durations);
those rows are set to `NA` and excluded from the daily sum.

Because `add_counts()` requires all design strata columns, merge the day type
back in from the raw timestamps:

```{r ie-merge-strata}
day_type_key <- unique(example_camera_timestamps[, c("date", "day_type")])
daily_effort <- merge(daily_effort, day_type_key, by = "date")
print(daily_effort)
```

### Build the Design and Estimate Effort

```{r ie-design}
ie_calendar <- data.frame(
  date = daily_effort$date,
  day_type = daily_effort$day_type,
  stringsAsFactors = FALSE
)

design_ie <- creel_design(
  ie_calendar,
  date        = date,
  strata      = day_type,
  survey_type = "camera",
  camera_mode = "ingress_egress"
)

design_ie <- add_counts(design_ie, daily_effort)
effort_ie <- suppressWarnings(estimate_effort(design_ie))
print(effort_ie)
```

## Interview-Based Catch Estimation

Camera designs support the same interview workflow as standard instantaneous
designs. Use the counter-mode design (which spans all eight interview dates)
and attach the interview data with `add_interviews()`.

```{r catch-interviews}
design_catch <- add_interviews(
  design_counter,
  example_camera_interviews,
  catch       = walleye,
  effort      = hours_fished,
  trip_status = trip_status
)
```

### Catch Rate

`estimate_catch_rate()` computes the ratio-of-means CPUE (walleye per
angler-hour) using only complete trips.

```{r catch-rate}
catch_rate <- suppressWarnings(estimate_catch_rate(design_catch))
print(catch_rate)
```

### Total Catch

`estimate_total_catch()` multiplies the CPUE estimate by the total effort
estimate to produce a projected total walleye catch over the survey period.

```{r total-catch}
total_catch <- suppressWarnings(estimate_total_catch(design_catch))
print(total_catch)
```

## Summary

The table below contrasts the two camera sub-modes:

```{r summary-table, echo = FALSE}
knitr::kable(data.frame(
  Feature = c(
    "Input data",
    "Preprocessing step",
    "Effort unit",
    "Gap handling",
    "camera_mode value"
  ),
  `Counter mode` = c(
    "One count per day",
    "None (counts used directly)",
    "Daily ingress count",
    "Exclude camera_status != 'operational' rows",
    '"counter"'
  ),
  `Ingress-egress mode` = c(
    "POSIXct arrival/departure pairs",
    "preprocess_camera_timestamps()",
    "Daily effort-hours",
    "Negative durations warned and excluded",
    '"ingress_egress"'
  ),
  check.names = FALSE
), caption = "Comparison of camera survey sub-modes in tidycreel")
```

Both sub-modes feed into the same `estimate_effort()`, `add_interviews()`,
`estimate_catch_rate()`, and `estimate_total_catch()` pipeline — no changes
to downstream code are required when switching modes.

## References

- Jones, C. M., & Pollock, K. H. (2012). Recreational survey methods:
  estimation of effort, harvest, and abundance. Chapter 19 in *Fisheries
  Techniques* (3rd ed.), pp. 883-919. American Fisheries Society.

- Malvestuto, S. P. (1996). Sampling the recreational angler. Chapter 20 in
  *Fisheries Techniques* (2nd ed.), pp. 591-623. American Fisheries Society.
