---
title: "Flexible Count Estimation"
output:
  rmarkdown::html_vignette:
    highlight: null
vignette: >
  %\VignetteIndexEntry{Flexible Count Estimation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Introduction

Creel surveys estimate angler effort — the total angler-hours or angler-trips on a waterbody
during a survey period. The count data collected in the field can take two forms:

- **Instantaneous counts** (Hoenig et al. 1993): At a random time within the survey period,
  an observer records the number of anglers visible on a section. Effort is estimated as
  count × total open hours.

- **Progressive counts** (Pope et al., in press): An observer travels the entire section
  (one "circuit") and records the number of anglers encountered along the way.
  Effort is estimated as count × circuit travel time × section open hours / circuit travel time,
  i.e., Ê_d = C × τ × (T_d / τ) = C × T_d where τ is circuit travel time and T_d is the
  total open hours on day d.

In addition, observers may complete **multiple counts per day** (e.g., morning and afternoon
circuits). When multiple counts are recorded per sampling unit (PSU), tidycreel aggregates
them automatically and decomposes variance into a between-day component and a within-day
component via the Rasmussen (1994) two-stage formula.

This vignette demonstrates all three workflows using inline data.

## Single Count per Day (Instantaneous Baseline)

The classic workflow records one instantaneous count per PSU (day × stratum).

```{r single-count, message = FALSE}
library(tidycreel)

# Survey calendar: four days, two strata
calendar <- data.frame(
  date = as.Date(c("2024-06-01", "2024-06-02", "2024-06-03", "2024-06-04")),
  day_type = c("weekday", "weekday", "weekend", "weekend"),
  open_hours = rep(10, 4)
)

design <- creel_design(calendar, date = date, strata = day_type)

# One count per day
counts <- data.frame(
  date      = as.Date(c("2024-06-01", "2024-06-02", "2024-06-03", "2024-06-04")),
  day_type  = c("weekday", "weekday", "weekend", "weekend"),
  n_anglers = c(15L, 23L, 45L, 52L)
)

design <- add_counts(design, counts)
result <- estimate_effort(design)
result$estimates
```

With a single count per day, `se_within` is always zero — there is no within-day
replication to estimate within-day variance. The `se` column equals `se_between`.

## Multiple Counts per Day

When observers complete multiple circuits on the same day, each circuit is a sub-sample
within the PSU. Supply the circuit identifier column via `count_time_col`.

```{r multi-count, message = FALSE}
# Two circuits per day: morning ("am") and afternoon ("pm")
calendar_m <- data.frame(
  date = as.Date(c("2024-06-01", "2024-06-02", "2024-06-03", "2024-06-04")),
  day_type = c("weekday", "weekday", "weekend", "weekend"),
  open_hours = rep(10, 4)
)

design_m <- creel_design(calendar_m, date = date, strata = day_type)

multi_counts <- data.frame(
  date = as.Date(rep(c(
    "2024-06-01", "2024-06-02",
    "2024-06-03", "2024-06-04"
  ), each = 2)),
  day_type = rep(c("weekday", "weekday", "weekend", "weekend"), each = 2),
  count_time = rep(c("am", "pm"), 4),
  n_anglers = c(12L, 18L, 20L, 26L, 40L, 50L, 48L, 56L)
)

design_m <- add_counts(design_m, multi_counts, count_time_col = count_time)
result_m <- estimate_effort(design_m)
result_m$estimates
```

The output tibble now contains nonzero `se_within` values. Total variance is
`se^2 = se_between^2 + se_within^2` (combined in quadrature). When within-day angler
counts vary substantially between circuits, `se_within` can be the dominant variance
component.

### When does `se_within` matter?

`se_within` is small when counts are stable throughout the day (e.g., a lake where
angler density is roughly constant). It is large when activity patterns are strongly
bimodal (e.g., a river with a morning bite and an evening bite but few mid-day anglers).
Collecting multiple counts per day is most valuable when such within-day variation is
expected and needs to be quantified rather than assumed away.

## Progressive Count Type

Progressive surveys replace a spot-check count with a complete traversal of the section.
The observer counts every angler encountered during one circuit of duration τ hours.
The estimator is:

$$\hat{E}_d = C_d \times T_d$$

where C_d is the count on day d and T_d is the total open hours on day d (the
`period_length_col` argument). The circuit time τ cancels algebraically; it is only
needed internally to compute the expansion factor κ = T_d / τ.

```{r progressive, message = FALSE}
calendar_p <- data.frame(
  date       = as.Date(c("2024-06-01", "2024-06-02", "2024-06-03", "2024-06-04")),
  day_type   = c("weekday", "weekday", "weekend", "weekend"),
  open_hours = rep(8, 4)
)

design_p <- creel_design(calendar_p, date = date, strata = day_type)

prog_counts <- data.frame(
  date = as.Date(c("2024-06-01", "2024-06-02", "2024-06-03", "2024-06-04")),
  day_type = c("weekday", "weekday", "weekend", "weekend"),
  n_anglers = c(15L, 23L, 45L, 52L),
  shift_hours = rep(8, 4)
)

design_p <- add_counts(
  design_p, prog_counts,
  count_type = "progressive",
  circuit_time = 2,
  period_length_col = shift_hours
)

result_p <- estimate_effort(design_p)
result_p$estimates
```

### Worked example (Pope et al.)

Pope et al. (in press) give a worked example with C = 234 anglers encountered during a
2-hour circuit (τ = 2 h) on a day with 8 open hours (T_d = 8 h):

$$\hat{E}_d = 234 \times 8 = 1{,}872 \text{ angler-hours}$$

The code below reproduces this per-day calculation for a two-day survey. The processed
design's `$counts` slot stores the expanded per-day effort (Ê_d), so we can confirm
the first-day value equals 1,872 angler-hours directly.

```{r pope-example, message = FALSE}
# Reproduce Pope et al. per-day calculation
cal_pope <- data.frame(
  date       = as.Date(c("2024-06-01", "2024-06-02")),
  day_type   = c("weekday", "weekday"),
  open_hours = c(8, 8)
)
design_pope <- creel_design(cal_pope, date = date, strata = day_type)

cnt_pope <- data.frame(
  date        = as.Date(c("2024-06-01", "2024-06-02")),
  day_type    = c("weekday", "weekday"),
  n_anglers   = c(234L, 200L),
  shift_hours = c(8, 8)
)

design_pope <- add_counts(
  design_pope, cnt_pope,
  count_type = "progressive",
  circuit_time = 2,
  period_length_col = shift_hours
)

# Per-day Ê_d values stored in the design (first day = 1,872 angler-hours)
design_pope$counts
```

The `n_anglers` column in `design_pope$counts` shows 1,872 for June 1st — matching
the Pope et al. formula exactly.

## Interpreting `se_between` and `se_within`

| Column | Source | Zero when |
|--------|--------|-----------|
| `se_between` | Between-PSU variance via `survey::svytotal()` | Never (always estimated) |
| `se_within` | Rasmussen (1994) two-stage within-day formula | Single count per PSU |
| `se` | Combined SE: `sqrt(se_between^2 + se_within^2)` | Never |

**Practical guidance:**

- If you collect one count per PSU, `se_within = 0` and `se = se_between`. The classical
  creel estimator is recovered exactly.

- If you collect multiple counts per PSU, monitor the ratio `se_within / se`. A large ratio
  (> 0.5) indicates substantial within-day variation and suggests that daily sampling
  intensity (number of circuits) is a meaningful driver of precision.

- For progressive surveys, the estimator variance structure is the same as for instantaneous
  counts. The key difference is in how the single-PSU effort Ê_d is calculated.

## References

- Hoenig, J. M., Robson, D. S., Jones, C. M., and Pollock, K. H. (1993). Scheduling counts
  in the instantaneous and progressive count methods for estimating sportfishing effort.
  *North American Journal of Fisheries Management*, 13, 723–736.

- Rasmussen, P. W. (1994). Two-stage variance estimation for creel surveys.
  Cited in Su, Y.-S., and Liu, P. (2025), equations 17–18.

- Su, Y.-S., and Liu, P. (2025). Flexible creel survey estimators.
  *Canadian Journal of Fisheries and Aquatic Sciences*, 82, 1–27.
