---
title: "Ice Fishing Survey Analysis"
output:
  rmarkdown::html_vignette:
    highlight: null
vignette: >
  %\VignetteIndexEntry{Ice Fishing Survey Analysis}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Overview

Ice fishing creel surveys share the same structure as bus-route surveys, but
with one important simplification: all access points are sampled with
certainty. Every angler leaving the lake must pass through the single ice
access point (the boat ramp area or designated ice-access site), so there is
no site-level subsampling. This means the site inclusion probability is always
`p_site = 1.0`, and the overall inclusion probability reduces to just the
period sampling probability:

$$\pi_i = p\_site \times p\_period = 1.0 \times p\_period = p\_period$$

Because of this, ice fishing surveys are implemented as a *degenerate*
bus-route design. All the bus-route Horvitz-Thompson estimators apply
unchanged — only the probability structure is simplified.

### Effort type distinction

Ice fishing surveys collect two distinct effort measures:

- **time_on_ice** — total hours the angler party was physically on the ice,
  including travel to the fishing hole, breaks, and social time. This is the
  easiest to record and most commonly used.
- **active_fishing_time** — hours spent with lines in the water, excluding
  travel, setup, and breaks. This captures actual fishing pressure more
  precisely but requires more careful interviewing.

The `effort_type` argument to `creel_design()` controls which measure the
design tracks and how the output column is labeled in `estimate_effort()`:
`total_effort_hr_on_ice` for `"time_on_ice"` and `total_effort_hr_active`
for `"active_fishing_time"`.

### Shelter mode stratification

Ice anglers fish from open-air setups or enclosed dark-house shelters. Catch
rates and effort patterns differ between these groups — dark-house anglers
often target specific species and fish longer hours. The `estimate_effort()`
function accepts a `by` argument to produce separate estimates for each
shelter type.

## Example Data

This vignette uses two built-in datasets representing a hypothetical Nebraska
ice fishing creel survey at Lake McConaughy in January-February 2024. All
12 sampling days are weekends — a common design choice since ice fishing
pressure is concentrated on Saturday and Sunday.

```{r data}
library(tidycreel)

data(example_ice_sampling_frame)
data(example_ice_interviews)

head(example_ice_sampling_frame)
head(example_ice_interviews)
```

The sampling frame records the survey calendar and the period sampling
probability (`p_period = 0.5` means each day had a 50% chance of being
sampled). The interview data contains 72 interviews with walleye and perch
catch, both effort measures, and the shelter type (`shelter_mode`) for each
angler party.

## Design Construction

Build an ice fishing design using `creel_design()` with
`survey_type = "ice"`. The `effort_type` argument is required and determines
the column label in downstream output. We use `p_period = 0.5` as a scalar
(uniform period sampling probability across all days).

```{r design}
design <- creel_design(
  example_ice_sampling_frame,
  date = date,
  strata = day_type,
  survey_type = "ice",
  effort_type = "time_on_ice",
  p_period = 0.5
)
print(design)
```

Omitting `effort_type` or supplying an unrecognized value produces an
informative error:

```{r design-error, error = TRUE}
creel_design(
  example_ice_sampling_frame,
  date = date,
  strata = day_type,
  survey_type = "ice"
)
```

## Attaching Interview Data

Use `add_interviews()` to attach the survey data to the design. For ice
surveys, `n_counted` and `n_interviewed` are required — they record how many
parties were counted at the access point versus how many were actually
interviewed during each visit, providing the enumeration expansion factor.

```{r interviews}
design <- add_interviews(
  design,
  example_ice_interviews,
  catch         = walleye_catch,
  effort        = hours_on_ice,
  harvest       = walleye_kept,
  trip_status   = trip_status,
  n_counted     = n_counted,
  n_interviewed = n_interviewed
)
```

## Effort Estimation

`estimate_effort()` dispatches through the bus-route Horvitz-Thompson
estimator and returns the total angler-hours with a standard error and
confidence interval. The output column is labeled `total_effort_hr_on_ice`
because the design was built with `effort_type = "time_on_ice"`.

```{r effort}
effort_est <- estimate_effort(design)
print(effort_est)
```

To demonstrate the column-naming distinction, rebuild the design using
`effort_type = "active_fishing_time"` and the `active_fishing_hours`
column from the interview data. The output column is then labeled
`total_effort_hr_active`.

```{r effort-active}
design_aft <- creel_design(
  example_ice_sampling_frame,
  date        = date,
  strata      = day_type,
  survey_type = "ice",
  effort_type = "active_fishing_time",
  p_period    = 0.5
)

design_aft <- suppressMessages(add_interviews(
  design_aft,
  example_ice_interviews,
  catch         = walleye_catch,
  effort        = active_fishing_hours,
  harvest       = walleye_kept,
  trip_status   = trip_status,
  n_counted     = n_counted,
  n_interviewed = n_interviewed
))

effort_aft <- estimate_effort(design_aft)
print(effort_aft)
```

Active fishing hours are shorter than time-on-ice because they exclude travel,
setup, and breaks. The difference between the two estimates reflects the
non-fishing portion of each trip.

## Shelter Mode Stratification

Pass `by = shelter_mode` to `estimate_effort()` to produce separate effort
estimates for open-air anglers and dark-house anglers. This uses the same
Horvitz-Thompson framework with the interview data split by the grouping
variable.

```{r effort-by-shelter}
effort_by_shelter <- estimate_effort(design, by = shelter_mode)
print(effort_by_shelter)
```

The `proportion` column shows each shelter group's share of total effort.
Dark-house anglers tend to fish longer hours on average, which can make their
proportional effort contribution larger than their share of party counts alone
would suggest.

## Catch Rate Estimation

`estimate_catch_rate()` computes the ratio-of-means CPUE (fish per
angler-hour) using the catch and effort columns specified in
`add_interviews()`. By default, only complete trips are used to avoid
the well-known incomplete-trip bias in effort-based catch rates.

```{r catch-rate}
cpue_est <- estimate_catch_rate(design)
print(cpue_est)
```

The `estimate` column is walleye per hour-on-ice, with a standard error and
95% confidence interval. The `n` column counts the number of complete
interviews used in the ratio.

## Total Catch Estimation

`estimate_total_catch()` multiplies the CPUE estimate by the total effort
estimate to produce a total fish catch estimate. It uses all interviews
(complete and incomplete) for the effort component and only complete trips
for the catch rate, following standard creel survey methodology.

```{r total-catch}
total_catch_est <- estimate_total_catch(design)
print(total_catch_est)
```

The `estimate` column is the projected total walleye catch across the entire
survey period, with a standard error and 95% confidence interval computed via
the delta method.

## Harvest Rate Estimation

`estimate_harvest_rate()` applies the Horvitz-Thompson estimator to the
harvest column (fish kept) and computes a total harvest rate over the survey
period. For ice designs the formula is:

$$\hat{H} = \sum_{i} \frac{h_i \times \text{expansion}_i}{\pi_i}$$

where $h_i$ is the fish-kept count for interview $i$, expansion is
`n_counted / n_interviewed`, and $\pi_i = p\_period$ (since `p_site = 1.0`
for ice surveys).

```{r harvest-rate}
harvest_rate_est <- estimate_harvest_rate(design)
print(harvest_rate_est)
```

## Total Harvest Estimation

`estimate_total_harvest()` uses the same HT formula applied directly to
expanded harvest counts, without requiring count survey data from
`add_counts()`. This is the key advantage of the bus-route HT approach:
effort and harvest come from the same interview rows.

```{r total-harvest}
total_harvest_est <- estimate_total_harvest(design)
print(total_harvest_est)
```

The `estimate` column is the projected total walleye harvest across all sampled
days, with standard error and 95% confidence interval. The result matches the
harvest rate estimate from `estimate_harvest_rate()` because both use the same
HT Eq. 19.5 applied to the harvest column.

## Total Release Estimation

To estimate total released fish, attach a catch data frame with
`catch_type = "released"` records using `add_catch()`. The function then
computes:

$$\hat{R} = \sum_{i} \frac{r_i \times \text{expansion}_i}{\pi_i}$$

where $r_i$ is the per-interview release count. The example data does not
include release records, but the workflow follows the standard `add_catch()`
pattern:

```{r total-release-setup, eval = FALSE}
design <- add_catch(
  design, catch_data,
  catch_uid = interview_id,
  interview_uid = interview_id,
  species = species,
  count = count,
  catch_type = catch_type
)
total_release_est <- estimate_total_release(design)
print(total_release_est)
```

## References

- Jones, C. M., & Pollock, K. H. (2012). Recreational survey methods:
  estimation of effort, harvest, and abundance. Chapter 19 in *Fisheries
  Techniques* (3rd ed.), pp. 883-919. American Fisheries Society.

- Malvestuto, S. P. (1996). Sampling the recreational angler. Chapter 20 in
  *Fisheries Techniques* (2nd ed.), pp. 591-623. American Fisheries Society.
