---
title: "Mark-Recapture and Exploitation Rate Estimation"
output:
  rmarkdown::html_vignette:
    highlight: null
vignette: >
  %\VignetteIndexEntry{Mark-Recapture and Exploitation Rate Estimation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Introduction

Creel surveys are often paired with tagging studies to estimate either angler
population size or seasonal exploitation rate. tidycreel provides three
estimators for these purposes:

- **`estimate_angler_n()`** — closed-population mark-recapture for total angler
  count (Chapman, Petersen, or Schnabel)
- **`estimate_mr_harvest()`** — total harvest derived from a mark-recapture
  population estimate
- **`estimate_exploitation_rate()`** — seasonal exploitation rate from
  tagged-fish recoveries (Pollock et al. 1994)

All three return `creel_estimates` objects compatible with the standard
tidycreel output ecosystem (`print()`, `autoplot()`, `write_estimates()`).

| Goal | Estimator |
|---|---|
| Estimate total anglers from a tag-and-resight study | `estimate_angler_n()` |
| Convert angler population estimate to total harvest | `estimate_mr_harvest()` |
| Estimate fraction of population harvested from tagged-fish recoveries | `estimate_exploitation_rate()` |

---

## Angler Population Size

### Chapman estimator (default)

The Chapman estimator is a bias-corrected Petersen estimator recommended
when the recapture count is small. With $M$ tagged anglers released, $n$
anglers checked in the second sample, and $m$ recaptures:

$$\hat{N} = \frac{(M+1)(n+1)}{(m+1)} - 1$$

Variance:

$$\widehat{\text{Var}}(\hat{N}) = \frac{(M+1)(n+1)(M-m)(n-m)}{(m+1)^2(m+2)}$$

```{r}
library(tidycreel)

result_chapman <- estimate_angler_n(M = 200L, n = 50L, m = 10L)
print(result_chapman)
```

The `n` column in the estimates tibble records the recapture count (`m`),
which determines precision.

### Petersen estimator

The unadjusted Lincoln–Petersen estimator:

$$\hat{N} = \frac{M \cdot n}{m}$$

tidycreel enforces a minimum of $m \geq 7$ recaptures; below this threshold
the Petersen estimator carries large positive bias and Chapman should be used
instead.

```{r}
result_petersen <- estimate_angler_n(M = 200L, n = 50L, m = 10L, method = "petersen")
print(result_petersen)
```

### Schnabel estimator (multi-occasion)

For $K \geq 2$ successive sampling occasions, the Schnabel estimator pools
information across occasions:

$$\hat{N} = \frac{\sum_{k=1}^{K} M_k n_k}{\sum_{k=1}^{K} m_k}$$

where $M_k$ is the cumulative count of marked-at-large anglers before
occasion $k$ (so $M_1 = 0$), $n_k$ is the catch on occasion $k$, and $m_k$
is the number of recaptures.

Confidence intervals use the Poisson distribution when $\sum m_k < 50$ and
a normal approximation on $1/\hat{N}$ otherwise.

```{r}
result_schnabel <- estimate_angler_n(
  M      = c(0L, 47L, 91L, 131L),
  n      = c(50L, 50L, 50L, 50L),
  m      = c(0L,  4L,  6L,  8L),
  method = "schnabel"
)
print(result_schnabel)
```

Four sampling occasions yield 18 total recaptures here, so the Poisson CI
branch is used automatically.

---

## Total Harvest from Mark-Recapture

Once angler population size is estimated, `estimate_mr_harvest()` scales
it by a known harvest rate to obtain total harvest. The harvest rate is
typically derived from creel interview data — the proportion of interviewed
anglers who kept fish.

$$\hat{H} = \hat{N} \times r, \quad SE(\hat{H}) = r \times SE(\hat{N})$$

```{r}
# harvest_rate derived from creel interviews: 35% of anglers harvested fish
harvest <- estimate_mr_harvest(angler_n = result_chapman, harvest_rate = 0.35)
print(harvest)
```

The delta method propagates uncertainty in $\hat{N}$ only. Uncertainty in
the harvest rate itself is not propagated in this release; if harvest rate
uncertainty is substantial, re-run across plausible bounds as a sensitivity
check.

---

## Exploitation Rate

`estimate_exploitation_rate()` implements the Pollock et al. (1994) moment
estimator. Rather than counting anglers, it estimates the fraction of a
tagged cohort that was harvested during the season. The inputs come from two
sources:

- **Tagging study**: $T$ fish tagged and released at season start; $m$ tagged
  fish recovered among $n$ fish inspected in the creel.
- **Creel survey**: $C$ total estimated harvest (e.g., from
  `estimate_total_catch()`) with standard error $SE_C$.

### Unstratified

$$\hat{u} = \frac{C \cdot m}{T \cdot n}, \quad
\widehat{\text{Var}}(\hat{u}) \approx
  \left(\frac{C}{T}\right)^2 \frac{p(1-p)}{n}
  + p^2 \frac{SE_C^2}{T^2}$$

where $p = m/n$.

```{r}
result_expl <- estimate_exploitation_rate(
  T    = 200L,
  C    = 450.0,
  se_C = 42.0,
  n    = 180L,
  m    = 15L
)
print(result_expl)
```

### Stratified (T-weighted)

When the survey spans multiple strata (day types, access areas), supply a
data frame with one row per stratum. The T-weighted aggregate is automatically
appended as an `.overall` row:

$$\hat{u}_{overall} = \frac{\sum_h T_h \hat{u}_h}{\sum_h T_h}, \quad
\widehat{\text{Var}}(\hat{u}_{overall}) =
  \frac{\sum_h T_h^2 \,\widehat{\text{Var}}(\hat{u}_h)}{(\sum_h T_h)^2}$$

```{r}
strata_df <- data.frame(
  stratum = c("weekday", "weekend"),
  T_h     = c(120L, 80L),
  C_h     = c(280.0, 170.0),
  se_C_h  = c(28.0, 22.0),
  n_h     = c(110L, 70L),
  m_h     = c(9L, 6L)
)

result_strat <- estimate_exploitation_rate(strata = strata_df, by = "stratum")
print(result_strat)
```

Pass `aggregate = FALSE` to suppress the `.overall` row.

### Reporting rate adjustment

If not all harvested tagged fish are reported, supply the reporting rate
$\lambda \in (0, 1]$:

$$\hat{u}_{adj} = \frac{\hat{u}}{\lambda}$$

```{r}
result_adj <- estimate_exploitation_rate(
  T              = 200L,
  C              = 450.0,
  se_C           = 42.0,
  n              = 180L,
  m              = 15L,
  reporting_rate = 0.80
)
print(result_adj)
```

Uncertainty in $\lambda$ is not propagated; treat reporting rate accuracy
as a separate sensitivity analysis.

---

## Choosing an Estimator

| Scenario | Recommended estimator |
|---|---|
| Single recapture event, small $m$ ($< 7$) | `estimate_angler_n(method = "chapman")` |
| Single recapture event, $m \geq 7$ | `"chapman"` or `"petersen"` (Chapman preferred) |
| Multiple survey occasions with cumulative marking | `estimate_angler_n(method = "schnabel")` |
| Converting population estimate to total harvest | `estimate_mr_harvest()` |
| Fraction of tagged cohort harvested (season-level) | `estimate_exploitation_rate()` |

---

## Assumptions

All three estimators share the standard closed-population assumptions:

1. **Closure** — the population is closed between marking and recapture; no
   births, deaths, emigration, or immigration.
2. **Equal catchability** — all individuals have equal probability of being
   captured or inspected on each occasion.
3. **Mark retention** — tags are not lost or overlooked at recapture.
4. **Random mixing** — marked individuals are randomly distributed in the
   population before the second sample.

For `estimate_exploitation_rate()` two additional caveats apply:

- Natural mortality between tagging and the creel survey is not corrected
  for; if natural mortality is non-trivial, the estimator overestimates
  exploitation.
- The reporting rate is treated as known without error (see adjustment
  section above).

---

## References

Hansen, M. J., & Van Kirk, R. W. (2018). A mark-recapture-based approach
for estimating angler harvest. *North American Journal of Fisheries
Management*, 38(2), 400–410. https://doi.org/10.1002/nafm.10038

Pollock, K. H., Jones, C. M., & Brown, T. L. (1994). *Angler Survey Methods
and Their Applications in Fisheries Management.* AFS Special Publication 25.
American Fisheries Society.

Jones, C. M., & Pollock, K. H. (2012). Recreational survey methods:
estimating effort, harvest, and abundance. In A. V. Zale et al. (Eds.),
*Fisheries Techniques* (3rd ed., Ch. 19). American Fisheries Society.

---

## See Also

- [Interview Estimation](interview-estimation.html) — deriving harvest rates
  from creel interviews
- [Survey Design Toolbox](survey-design-toolbox.html) — power and sample-size
  planning
