---
title: "Variance Methods: Taylor, Bootstrap, and Jackknife"
output:
  rmarkdown::html_vignette:
    highlight: null
vignette: >
  %\VignetteIndexEntry{Variance Methods: Taylor, Bootstrap, and Jackknife}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4,
  out.width = "100%"
)
```

tidycreel exposes three variance estimation methods through the `variance`
argument of `estimate_effort()`, `estimate_catch_rate()`, and related
functions:

| `variance =` | Description | Requires |
|---|---|---|
| `"taylor"` | Taylor linearization (default) | ≥ 2 PSUs per stratum recommended |
| `"bootstrap"` | Bootstrap resampling (500 replicates) | ≥ 2 PSUs per stratum |
| `"jackknife"` | Delete-one jackknife | ≥ 2 PSUs per stratum |

```{r load}
library(tidycreel)
```

---

## 1  Building a design for the examples

All three methods work on the same `creel_design` object. Here we use the
bundled `example_counts` and `example_interviews` datasets, which have
10 sampled weekdays and 4 sampled weekends.

```{r build-design, warning = FALSE}
data("example_counts")
data("example_interviews")

cal <- unique(example_counts[, c("date", "day_type")])
design <- creel_design(cal, date = date, strata = day_type)
design <- add_counts(design, example_counts)
design <- add_interviews(
  design, example_interviews,
  catch = catch_total,
  effort = hours_fished,
  trip_status = trip_status
)
```

---

## 2  Comparing the three variance methods

Pass `variance = "..."` to any estimator. The point estimate is identical
across methods; only the SE and CI change.

```{r compare, warning = FALSE}
eff_taylor <- estimate_effort(design, variance = "taylor")
eff_bootstrap <- estimate_effort(design, variance = "bootstrap")
eff_jackknife <- estimate_effort(design, variance = "jackknife")

# Summarise all three for comparison
rbind(
  as.data.frame(summary(eff_taylor)),
  as.data.frame(summary(eff_bootstrap)),
  as.data.frame(summary(eff_jackknife))
)
```

With 10–14 PSUs per stratum the three methods produce very similar standard
errors. This convergence is expected: bootstrap and jackknife are
asymptotically equivalent to Taylor linearization for simple probability
samples.

---

## 3  When to use each method

### Taylor linearization (`"taylor"`, default)

- **Use when**: stratum-level variance is dominated by between-day
  variability and the sampling design is a straightforward stratified random
  sample of days.
- **Advantage**: fastest; closed-form analytic variance formula; no
  randomness.
- **Limitation**: relies on large-sample approximations; can be less reliable
  for highly skewed count distributions or very small strata.

### Bootstrap (`"bootstrap"`)

- **Use when**: count distributions are highly skewed, the design is complex,
  or you want variance that is robust to distributional assumptions.
- **Advantage**: makes no distributional assumption; captures nonlinear
  variance propagation.
- **Limitation**: stochastic (set a seed in your script for reproducibility);
  slower (500 replicates by default); requires ≥ 2 PSUs per stratum.

### Jackknife (`"jackknife"`)

- **Use when**: you prefer a deterministic, delete-one variance estimate that
  is slightly more conservative than Taylor for small samples.
- **Advantage**: deterministic; often more stable than bootstrap with small
  PSU counts.
- **Limitation**: also requires ≥ 2 PSUs per stratum; not appropriate for
  designs with unequal inclusion probabilities unless the correct jackknife
  weights are applied.

---

## 4  Grouped estimates

All three methods work with the `by` argument.

```{r grouped, warning = FALSE}
eff_grp_taylor <- estimate_effort(design, by = day_type, variance = "taylor")
eff_grp_bootstrap <- estimate_effort(design, by = day_type, variance = "bootstrap")

rbind(
  cbind(method = "taylor", as.data.frame(summary(eff_grp_taylor))),
  cbind(method = "bootstrap", as.data.frame(summary(eff_grp_bootstrap)))
)
```

---

## 5  Single-PSU diagnostic

All three variance methods require at least **2 PSUs (sampled days) per
stratum**. When a stratum has only 1 sampled day, tidycreel raises a
structured error naming the stratum:

```{r single-psu, error = TRUE, warning = FALSE}
# Build a minimal design with 1 sampled day per stratum
dates_1psu <- c(as.Date("2024-06-03"), as.Date("2024-06-08"))
cal_1psu <- data.frame(date = dates_1psu, day_type = c("weekday", "weekend"))
d_1psu <- creel_design(cal_1psu, date = date, strata = day_type)
counts_1psu <- data.frame(
  date     = dates_1psu,
  day_type = c("weekday", "weekend"),
  count    = c(10L, 15L)
)
d_1psu <- add_counts(d_1psu, counts_1psu)

# Taylor linearization raises an error when variance cannot be estimated
estimate_effort(d_1psu, variance = "jackknife")
```

The error message names the problematic stratum (`"weekday"`) and suggests
either increasing the sampling rate or switching to `variance = "taylor"`
which can still produce a point estimate via the lonely-PSU adjustment.

---

## 6  Summary

```{r summary-table, echo = FALSE}
knitr::kable(
  data.frame(
    Method = c("taylor", "bootstrap", "jackknife"),
    Deterministic = c("Yes", "No (stochastic)", "Yes"),
    `Min PSUs` = c("1 (with lonely-PSU adjust)", "2", "2"),
    Speed = c("Fast", "Slow (500 replicates)", "Moderate"),
    `When to prefer` = c(
      "Default; simple designs",
      "Skewed counts; complex designs",
      "Small samples; deterministic alternative to bootstrap"
    ),
    check.names = FALSE
  ),
  caption = "Variance method comparison"
)
```
