---
title: "Survey Scheduling"
output:
  rmarkdown::html_vignette:
    highlight: null
vignette: >
  %\VignetteIndexEntry{Survey Scheduling}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Overview

A well-structured sampling schedule ensures that creel survey estimates of effort, catch,
and CPUE are unbiased and representative of the full survey season. `generate_schedule()`
builds a stratified random sampling calendar from season boundaries and a target sampling
intensity — producing the count-period framework that drives the tidycreel estimation
pipeline.

## The Count Period Framework

Every tidycreel design rests on a calendar of **count periods** — the discrete time slots
during which field crews collect data. A count period is defined by:

- **Date**: the survey date
- **Period**: the time-of-day window (e.g., "Morning", "Afternoon")
- **Stratum**: the day type (weekday vs. weekend)

Interviews and angler counts are both collected during count periods. The schedule you
generate defines exactly which days and periods will be sampled.

## Building a Basic Schedule

The minimal inputs are season boundaries, the number of periods per day, and sampling
intensity. The example below builds a schedule for a May–September creel season with
morning and afternoon count periods, sampling 30% of weekdays and 50% of weekends.

```{r basic-schedule, message = FALSE}
library(tidycreel)

sched <- generate_schedule(
  start_date    = "2024-05-01",
  end_date      = "2024-09-30",
  n_periods     = 2,
  period_labels = c("Morning", "Afternoon"),
  sampling_rate = c(weekday = 0.30, weekend = 0.50),
  seed          = 42
)

head(sched, 8)
```

The result is a `creel_schedule` object with `date`, `day_type`, and `period_id` columns.
Period labels replace integer IDs with human-readable names.

```{r schedule-summary}
# Counts of scheduled days by stratum
table(sched$day_type)
```

## Visualising Sampled vs. Unsampled Days

Pass `include_all = TRUE` to return every day of the season with a `sampled` indicator.
This is useful for checking seasonal coverage and communicating the design to stakeholders.

```{r include-all, message = FALSE}
sched_full <- generate_schedule(
  start_date    = "2024-05-01",
  end_date      = "2024-09-30",
  n_periods     = 1,
  sampling_rate = c(weekday = 0.30, weekend = 0.50),
  include_all   = TRUE,
  seed          = 42
)

# Sampled proportion by stratum
table(sched_full$day_type, sched_full$sampled)
```

The sampled days are drawn by stratified random sampling within each stratum, so weekday
and weekend sampling fractions match the requested rates.

## Three-Period Schedules

Some surveys divide the day into three periods — morning, midday, and evening — to capture
temporal patterns in angler activity (e.g., a morning and evening peak with low midday use).

```{r three-periods, message = FALSE}
sched3 <- generate_schedule(
  start_date    = "2024-06-01",
  end_date      = "2024-08-31",
  n_periods     = 3,
  period_labels = c("Morning", "Midday", "Evening"),
  sampling_rate = c(weekday = 0.40, weekend = 0.60),
  seed          = 101
)

head(sched3, 9)
```

## Reproducibility with `seed`

The `seed` argument ensures that the same calendar is produced every time. This is
important when the schedule is pre-printed for field use and must match what the
estimators will use.

```{r seed}
sched_a <- generate_schedule("2024-06-01", "2024-06-30",
  n_periods = 1,
  sampling_rate = 0.4, seed = 99
)
sched_b <- generate_schedule("2024-06-01", "2024-06-30",
  n_periods = 1,
  sampling_rate = 0.4, seed = 99
)

identical(sched_a, sched_b)
```

## Saving and Sharing the Schedule

`write_schedule()` exports to CSV (for any spreadsheet application) or xlsx (for direct
field printing). `read_schedule()` reloads the file with correct column types, so the
schedule round-trips without type coercion issues.

```{r io, eval = FALSE}
# Save for field crews
write_schedule(sched, "count_schedule_2024.csv")
write_schedule(sched, "count_schedule_2024.xlsx")

# Reload — column types are preserved automatically
sched_reload <- read_schedule("count_schedule_2024.csv")
identical(sched, sched_reload)
```

## Linking the Schedule to a Design

Pass the schedule directly to `creel_design()`. The schedule's stratum structure informs
the survey weights used by all downstream estimators.

```{r design, message = FALSE}
design <- creel_design(sched, date = date, strata = day_type)
design
```

## Count Period Scheduling vs. Interview Scheduling

In standard access-point creel surveys, **interviews happen during count periods** — the
same field crew that counts anglers also approaches them for interviews. The count schedule
therefore governs both collection activities.

There is no separate "interview schedule." The set of available interviews is the set of
interviews collected on sampled count-period days.

## Bus-Route Scheduling

For bus-route designs, `generate_bus_schedule()` converts a count schedule and a
site-level sampling frame into a complete bus-route sampling frame with inclusion
probabilities. Each site's probability of being visited in a given period is:

$$\pi_i = p_{site} \times p_{period} \quad \text{where} \quad p_{period} = \frac{\text{crew}}{n_{circuits}}$$

```{r bus-schedule, message = FALSE}
# Site-level selection probabilities (must sum to 1.0)
site_frame <- data.frame(
  site_id = c("North Bay", "South Cove", "Dock Area", "Main Channel"),
  p_site  = c(0.35, 0.25, 0.25, 0.15)
)

# Build bus-route sampling frame using the count schedule
bus_frame <- generate_bus_schedule(
  schedule       = sched,
  sampling_frame = site_frame,
  site           = site_id,
  p_site         = p_site,
  crew           = 1
)

bus_frame
```

The `inclusion_prob` column is the per-site overall inclusion probability across the
season. This frame passes directly to `creel_design(survey_type = "bus_route", ...)`.

## Sampling Rate Guidance

The right sampling intensity depends on the target precision for the survey. As a
starting point:

| Survey length | Typical weekday rate | Typical weekend rate |
|---------------|---------------------|---------------------|
| Short (< 8 weeks) | 0.40–0.60 | 0.60–0.80 |
| Standard (8–20 weeks) | 0.25–0.40 | 0.40–0.60 |
| Long (> 20 weeks) | 0.15–0.25 | 0.25–0.40 |

Higher rates on weekends reflect greater angler activity and higher stratum variance.
Use `creel_n_effort()` or `creel_power()` to select rates based on a CV target rather
than rule of thumb.

```{r sample-size, message = FALSE}
# Effort sample size to achieve CV ≤ 0.15
# N_h: total days per stratum in a 184-day season
# ybar_h: pilot mean daily effort (angler-hours) per stratum
# s2_h: pilot variance of daily effort per stratum
creel_n_effort(
  cv_target = 0.15,
  N_h = c(weekday = 132, weekend = 52),
  ybar_h = c(weekday = 280, weekend = 550),
  s2_h = c(weekday = 14400, weekend = 32400)
)
```

## Within-Day Count Time Scheduling

Counts are the primary effort estimator in creel surveys: the number of anglers observed
during each count window is used to estimate total angler-hours. When those count windows
fall in the day matters — clustered or poorly spaced windows can introduce bias into effort
estimates. Pollock et al. (1994) describe three strategies for placing count time windows
within the survey day: random, systematic, and fixed. `generate_count_times()` implements
all three.

### Random Strategy

With the random strategy, `n_windows` non-overlapping windows are drawn uniformly at
random from the available survey hours. `window_size` sets the duration (minutes) of each
window and `min_gap` enforces a minimum separation (minutes) between consecutive windows
to prevent back-to-back counts.

```{r count-times-random, message = FALSE}
library(tidycreel)
ct_random <- generate_count_times(
  start_time  = "06:00",
  end_time    = "14:00",
  strategy    = "random",
  n_windows   = 4,
  window_size = 30,
  min_gap     = 10,
  seed        = 42
)
ct_random
```

`n_windows` is the total number of count windows to place in the day, `window_size` is
how long each window lasts in minutes, and `min_gap` is the minimum idle time between the
end of one window and the start of the next.

### Systematic Strategy

The systematic strategy uses the Pollock et al. (1994) model: the day is divided into
`n_windows` equal-length strata and one window is placed at a random start within the
first stratum; subsequent windows begin at t₁+k, t₁+2k, and so on, where k is the
stratum length. This ensures even spacing across the day. Systematic sampling is
preferred by many agencies (Colorado CPW 2012) because it avoids the clustered count
times that can occur by chance under random placement.

```{r count-times-systematic, message = FALSE}
ct_systematic <- generate_count_times(
  start_time  = "06:00",
  end_time    = "14:00",
  strategy    = "systematic",
  n_windows   = 4,
  window_size = 30,
  min_gap     = 10,
  seed        = 42
)
ct_systematic
```

### Fixed Strategy

Use the fixed strategy when count time windows are defined by regulation, permit
conditions, or a prior-season protocol that must be replicated exactly. Supply a
`data.frame` with `start_time` and `end_time` columns and `generate_count_times()` wraps
them in a `creel_schedule` object without any random placement.

```{r count-times-fixed, message = FALSE}
fw <- data.frame(
  start_time = c("07:00", "09:30", "12:00"),
  end_time = c("07:30", "10:00", "12:30"),
  stringsAsFactors = FALSE
)
ct_fixed <- generate_count_times(strategy = "fixed", fixed_windows = fw)
ct_fixed
```

### Exporting Count Time Schedules

`generate_count_times()` returns a `creel_schedule` object, the same class returned by
`generate_schedule()`. It therefore passes directly to `write_schedule()` for field
printing without any conversion step.

```{r count-times-export, eval = FALSE}
write_schedule(ct_systematic, "count_times_2024.csv")
```

## Combining the Daily Schedule with Count Time Windows

`generate_schedule()` and `generate_count_times()` each return a separate `creel_schedule`.
Before field dispatch, biologists need a single table — one row per (date x period x count
window) — that tells each crew exactly when to count. `attach_count_times()` performs the
cross-join in one step.

```{r attach-count-times, message = FALSE}
sched <- generate_schedule(
  start_date = "2024-06-01", end_date = "2024-06-07",
  n_periods = 2,
  sampling_rate = c(weekday = 0.5, weekend = 0.8),
  seed = 42
)
ct <- generate_count_times(
  start_time = "06:00", end_time = "14:00",
  strategy = "systematic", n_windows = 3,
  window_size = 30, min_gap = 10,
  seed = 42
)
field_schedule <- attach_count_times(sched, ct)
field_schedule
```

The result has `nrow(sched) * nrow(ct)` rows — every sampled day-period combination
appears once per count window. Pass `field_schedule` directly to `write_schedule()` to
produce a printable field dispatch sheet.

## Validating the Design Before the Season

After building a schedule, `validate_design()` checks whether the proposed sampling
intensity is sufficient to meet a target coefficient of variation (CV) — catching
under-sampling problems before the season starts rather than discovering them in the
post-season analysis. It uses the same pilot mean and variance inputs as `creel_n_effort()`
and compares the proposed day counts against the minimum required sample size per stratum.

```{r validate-design, message = FALSE}
report <- validate_design(
  N_h        = c(weekday = 132, weekend = 52),
  ybar_h     = c(weekday = 280, weekend = 550),
  s2_h       = c(weekday = 14400, weekend = 32400),
  n_proposed = c(weekday = 40L, weekend = 26L),
  cv_target  = 0.15
)
report
```

`report$results` shows the per-stratum status (pass / warn / fail), the proposed sample
size, the required sample size, and the achieved CV at the proposed intensity.
`report$passed` is `TRUE` only when all strata pass — use this as a go/no-go gate before
printing field schedules.

```{r validate-design-results, message = FALSE}
report$results
```

## Checking Data Completeness After the Season

`check_completeness()` runs post-season on a `creel_design` with survey data attached,
flagging missing sampling days and strata with too few interviews to produce reliable
estimates. It is the first diagnostic step before running estimators — resolving
completeness issues early avoids propagating gaps into effort or CPUE estimates.

```{r check-completeness, message = FALSE}
data(example_calendar)
data(example_counts)
data(example_interviews)

design <- creel_design(example_calendar, date = date, strata = day_type)
design <- add_counts(design, example_counts)
design <- add_interviews(design, example_interviews,
  catch        = catch_total,
  effort       = hours_fished,
  trip_status  = trip_status
)
comp <- check_completeness(design)
comp
```

`comp$passed` returns `TRUE` if no issues were found. When `FALSE`, `comp$missing_days`
identifies dates that appear in the schedule but have no count data, and
`comp$low_n_strata` identifies strata where the interview count is below the minimum
threshold (`n_min`, default 10).

## Assembling the Season Summary

`season_summary()` accepts a named list of `creel_estimates` objects — the outputs of
`estimate_effort()`, `estimate_catch_rate()`, and related functions — and joins them into
a single wide tibble for reporting or export. It performs no re-estimation; it is purely
an assembly step that makes it easy to view all key metrics side by side.

```{r season-summary, eval = FALSE}
# Run estimators first (see the tidycreel workflow vignette)
effort <- estimate_effort(design)
cpue <- estimate_catch_rate(design)

# Assemble the season summary
summary_tbl <- season_summary(list(effort = effort, cpue = cpue))
summary_tbl$table
```

The full estimation workflow — building the design, attaching counts and interviews, and
running the estimators — is covered in the main tidycreel vignette. `season_summary()` is
the final step that wraps pre-computed results for export or reporting.

```{r season-summary-export, eval = FALSE}
write_schedule(summary_tbl$table, "season_2024_summary.xlsx")
```

`write_schedule()` accepts any data frame or tibble, so the season summary table exports
to CSV or xlsx with a single call.

## References

- Hoenig, J. M., Robson, D. S., Jones, C. M., and Pollock, K. H. (1993). Scheduling
  counts in the instantaneous and progressive count methods for estimating sportfishing
  effort. *North American Journal of Fisheries Management*, 13, 723–736.

- Pollock, K. H., Jones, C. M., and Brown, T. L. (1994). *Angler Survey Methods and Their
  Applications in Fisheries Management*. American Fisheries Society, Bethesda, MD.
