---
title: "Survey Scheduling: Count and Interview Periods"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Survey Scheduling}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Overview

A well-structured sampling schedule ensures that creel survey estimates of effort, catch,
and CPUE are unbiased and representative of the full survey season. `generate_schedule()`
builds a stratified random sampling calendar from season boundaries and a target sampling
intensity — producing the count-period framework that drives the tidycreel estimation
pipeline.

## The Count Period Framework

Every tidycreel design rests on a calendar of **count periods** — the discrete time slots
during which field crews collect data. A count period is defined by:

- **Date**: the survey date
- **Period**: the time-of-day window (e.g., "Morning", "Afternoon")
- **Stratum**: the day type (weekday vs. weekend)

Interviews and angler counts are both collected during count periods. The schedule you
generate defines exactly which days and periods will be sampled.

## Building a Basic Schedule

The minimal inputs are season boundaries, the number of periods per day, and sampling
intensity. The example below builds a schedule for a May–September creel season with
morning and afternoon count periods, sampling 30% of weekdays and 50% of weekends.

```{r basic-schedule, message = FALSE}
library(tidycreel)

sched <- generate_schedule(
  start_date    = "2024-05-01",
  end_date      = "2024-09-30",
  n_periods     = 2,
  period_labels = c("Morning", "Afternoon"),
  sampling_rate = c(weekday = 0.30, weekend = 0.50),
  seed          = 42
)

head(sched, 8)
```

The result is a `creel_schedule` object with `date`, `day_type`, and `period_id` columns.
Period labels replace integer IDs with human-readable names.

```{r schedule-summary}
# Counts of scheduled days by stratum
table(sched$day_type)
```

## Visualising Sampled vs. Unsampled Days

Pass `include_all = TRUE` to return every day of the season with a `sampled` indicator.
This is useful for checking seasonal coverage and communicating the design to stakeholders.

```{r include-all, message = FALSE}
sched_full <- generate_schedule(
  start_date    = "2024-05-01",
  end_date      = "2024-09-30",
  n_periods     = 1,
  sampling_rate = c(weekday = 0.30, weekend = 0.50),
  include_all   = TRUE,
  seed          = 42
)

# Sampled proportion by stratum
table(sched_full$day_type, sched_full$sampled)
```

The sampled days are drawn by stratified random sampling within each stratum, so weekday
and weekend sampling fractions match the requested rates.

## Three-Period Schedules

Some surveys divide the day into three periods — morning, midday, and evening — to capture
temporal patterns in angler activity (e.g., a morning and evening peak with low midday use).

```{r three-periods, message = FALSE}
sched3 <- generate_schedule(
  start_date    = "2024-06-01",
  end_date      = "2024-08-31",
  n_periods     = 3,
  period_labels = c("Morning", "Midday", "Evening"),
  sampling_rate = c(weekday = 0.40, weekend = 0.60),
  seed          = 101
)

head(sched3, 9)
```

## Reproducibility with `seed`

The `seed` argument ensures that the same calendar is produced every time. This is
important when the schedule is pre-printed for field use and must match what the
estimators will use.

```{r seed}
sched_a <- generate_schedule("2024-06-01", "2024-06-30",
  n_periods = 1,
  sampling_rate = 0.4, seed = 99
)
sched_b <- generate_schedule("2024-06-01", "2024-06-30",
  n_periods = 1,
  sampling_rate = 0.4, seed = 99
)

identical(sched_a, sched_b)
```

## Saving and Sharing the Schedule

`write_schedule()` exports to CSV (for any spreadsheet application) or xlsx (for direct
field printing). `read_schedule()` reloads the file with correct column types, so the
schedule round-trips without type coercion issues.

```{r io, eval = FALSE}
# Save for field crews
write_schedule(sched, "count_schedule_2024.csv")
write_schedule(sched, "count_schedule_2024.xlsx")

# Reload — column types are preserved automatically
sched_reload <- read_schedule("count_schedule_2024.csv")
identical(sched, sched_reload)
```

## Linking the Schedule to a Design

Pass the schedule directly to `creel_design()`. The schedule's stratum structure informs
the survey weights used by all downstream estimators.

```{r design, message = FALSE}
design <- creel_design(sched, date = date, strata = day_type)
design
```

## Count Period Scheduling vs. Interview Scheduling

In standard access-point creel surveys, **interviews happen during count periods** — the
same field crew that counts anglers also approaches them for interviews. The count schedule
therefore governs both collection activities.

There is no separate "interview schedule." The set of available interviews is the set of
interviews collected on sampled count-period days.

## Bus-Route Scheduling

For bus-route designs, `generate_bus_schedule()` converts a count schedule and a
site-level sampling frame into a complete bus-route sampling frame with inclusion
probabilities. Each site's probability of being visited in a given period is:

$$\pi_i = p_{site} \times p_{period} \quad \text{where} \quad p_{period} = \frac{\text{crew}}{n_{circuits}}$$

```{r bus-schedule, message = FALSE}
# Site-level selection probabilities (must sum to 1.0)
site_frame <- data.frame(
  site_id = c("North Bay", "South Cove", "Dock Area", "Main Channel"),
  p_site  = c(0.35, 0.25, 0.25, 0.15)
)

# Build bus-route sampling frame using the count schedule
bus_frame <- generate_bus_schedule(
  schedule       = sched,
  sampling_frame = site_frame,
  site           = site_id,
  p_site         = p_site,
  crew           = 1
)

bus_frame
```

The `inclusion_prob` column is the per-site overall inclusion probability across the
season. This frame passes directly to `creel_design(survey_type = "bus_route", ...)`.

## Sampling Rate Guidance

The right sampling intensity depends on the target precision for the survey. As a
starting point:

| Survey length | Typical weekday rate | Typical weekend rate |
|---------------|---------------------|---------------------|
| Short (< 8 weeks) | 0.40–0.60 | 0.60–0.80 |
| Standard (8–20 weeks) | 0.25–0.40 | 0.40–0.60 |
| Long (> 20 weeks) | 0.15–0.25 | 0.25–0.40 |

Higher rates on weekends reflect greater angler activity and higher stratum variance.
Use `creel_n_effort()` or `creel_power()` to select rates based on a CV target rather
than rule of thumb.

```{r sample-size, message = FALSE}
# Effort sample size to achieve CV ≤ 0.15
creel_n_effort(cv = 0.15, mean_effort = 80, sd_effort = 45)
```

## References

- Hoenig, J. M., Robson, D. S., Jones, C. M., and Pollock, K. H. (1993). Scheduling
  counts in the instantaneous and progressive count methods for estimating sportfishing
  effort. *North American Journal of Fisheries Management*, 13, 723–736.

- Pollock, K. H., Jones, C. M., and Brown, T. L. (1994). *Angler Survey Methods and Their
  Applications in Fisheries Management*. American Fisheries Society, Bethesda, MD.
