---
title: "Survey Design Toolbox: Planning, Comparing, and Combining Designs"
output:
  rmarkdown::html_vignette:
    highlight: null
vignette: >
  %\VignetteIndexEntry{Survey Design Toolbox: Planning, Comparing, and Combining Designs}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4,
  out.width = "100%"
)
```

This vignette collects three M014 tools into one practitioner-facing workflow:

1. `power_creel()` for pre-season sample-size and power planning
2. `compare_designs()` for side-by-side comparison of completed survey estimates
3. `as_hybrid_svydesign()` for combining access-point and roving count data in a
   single survey design

```{r load}
library(tidycreel)
```

## 1 Pre-season sample-size planning with `power_creel()`

Pre-season planning usually starts with pilot information: expected effort by
stratum, variability in daily counts, and rough interview-level CV values for
catch and effort. `power_creel()` provides one interface for three planning
questions.

### Required sampling days for total effort precision

This first example estimates how many sampling days are needed in weekday and
weekend strata to target a 20% RSE on the seasonal effort estimate.

```{r power-effort}
effort_plan <- power_creel(
  mode = "effort_n",
  target_rse = 0.20,
  strata = c("weekday", "weekend"),
  N_h = c(90, 30),
  ybar_h = c(42, 68),
  s2_h = c(196, 441)
)

effort_plan
```

The result returns one row per stratum plus a `total` row, making it easy to
translate a seasonal precision target into a day-allocation plan.

### Required interviews for CPUE precision

If the planning question is interview effort rather than count days,
`mode = "cpue_n"` solves for the number of interviews needed to estimate CPUE
with a target RSE.

```{r power-cpue}
cpue_plan <- power_creel(
  mode = "cpue_n",
  target_rse = 0.15,
  cv_catch = 0.85,
  cv_effort = 0.55,
  rho = 0.35
)

cpue_plan
```

This is useful when interview staffing is the main operational bottleneck and
pilot data already suggest the variability of catch and angler effort.

### Power to detect a management-relevant CPUE change

The third mode asks a different question: if we can complete a fixed number of
interviews, how much power do we have to detect a change in CPUE from one season
to the next?

```{r power-detect}
power_plan <- power_creel(
  mode = "power",
  n = 120L,
  cv_historical = 0.42,
  delta_pct = 0.20
)

power_plan
```

Here `delta_pct = 0.20` means a 20% change in CPUE. Together, the three modes
cover the most common pre-season planning decisions: how many days to sample,
how many interviews to complete, and what power that design can deliver.

## 2 Comparing finished designs with `compare_designs()`

Once a survey has been completed, `compare_designs()` helps compare multiple
`creel_estimates` objects on a common scale. In this example we estimate total
effort twice from the same dataset, changing only the variance method.

```{r compare-setup, warning = FALSE, message = FALSE}
data("example_counts")
data("example_interviews")

calendar <- unique(example_counts[, c("date", "day_type")])

design <- creel_design(calendar, date = date, strata = day_type)
design <- add_counts(design, example_counts)
design <- add_interviews(
  design,
  example_interviews,
  catch = catch_total,
  effort = hours_fished,
  trip_status = trip_status,
  n_anglers = n_anglers
)
```

```{r compare-estimates, warning = FALSE}
set.seed(123)

effort_taylor <- estimate_effort(design, variance = "taylor")
effort_bootstrap <- estimate_effort(design, variance = "bootstrap")

design_comparison <- compare_designs(
  list(
    Taylor = effort_taylor,
    Bootstrap = effort_bootstrap
  )
)

design_comparison
```

Because both estimates come from the same counts and interviews, the point
estimate is identical while the uncertainty metrics reflect the different
variance estimators.

```{r compare-plot, fig.cap = "Design comparison across two effort estimators."}
ggplot2::autoplot(design_comparison)
```

This pattern is helpful after a season when you want to compare alternative
estimation choices without rebuilding custom summary tables by hand.

## 3 Combining access and roving counts with `as_hybrid_svydesign()`

Some programs collect counts from both fixed access points and roving routes.
`as_hybrid_svydesign()` stacks those components into a single `survey` design
object with component-specific weights.

```{r hybrid-design}
access <- data.frame(
  date = as.Date(c("2024-06-03", "2024-06-08", "2024-06-10", "2024-06-15")),
  day_type = c("weekday", "weekend", "weekday", "weekend"),
  count = c(12L, 18L, 9L, 21L)
)

roving <- data.frame(
  date = as.Date(c("2024-06-03", "2024-06-08", "2024-06-10", "2024-06-15")),
  day_type = c("weekday", "weekend", "weekday", "weekend"),
  count = c(10L, 16L, 8L, 19L)
)

hybrid_design <- as_hybrid_svydesign(
  access_data = access,
  roving_data = roving,
  access_fraction = c(weekday = 0.5, weekend = 0.5),
  roving_fraction = c(weekday = 0.5, weekend = 0.5)
)

hybrid_design
```

With the combined design in hand, standard `survey` estimators work directly on
the stacked data.

```{r hybrid-total}
survey::svytotal(~count, hybrid_design)
```

This small example is intentionally self-contained, but the same pattern scales
to real field programs where access and roving counts represent complementary
views of the same sampling frame.

## Summary

The survey-design toolbox supports the full planning-to-reporting arc:

- `power_creel()` helps set realistic pre-season sample sizes and power targets
- `compare_designs()` turns alternative estimator outputs into a tidy, directly
  comparable object with a plotting method
- `as_hybrid_svydesign()` bridges mixed access-point and roving count programs
  into one survey design for downstream analysis

Used together, these tools make it easier to justify sampling effort before the
season, evaluate estimator trade-offs afterward, and support hybrid monitoring
programs with standard survey workflows.
