---
title: "Temporal Extrapolation: Monthly, Seasonal, and Annual Estimates"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Temporal Extrapolation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Overview

Creel surveys sample a fraction of days in a season. Design-based estimators in
tidycreel automatically account for unsampled days through survey weights — so when
you call `estimate_effort()` or `estimate_total_catch()`, the result is already a
season-total estimate, not just a sample-day sum.

This vignette shows how to:

1. Obtain a full-season total from a single design
2. Break the season into monthly totals using separate monthly designs
3. Combine monthly estimates to a season or annual total
4. Assemble multi-estimate reports with `season_summary()`

## How Design-Based Extrapolation Works

When you create a design with `creel_design()`, the survey object assigns a **weight**
to each sampled day. A day sampled at rate $f$ receives weight $1/f$, so its
observed angler-hours represent $1/f$ days of that stratum type. The estimator
then sums weighted observations across strata to produce a population total for
the entire design period.

In practice this means:

- A season design covering May–September gives you a **season-total** estimate.
- A monthly design covering just June gives you a **June-total** estimate.
- You do not need to multiply results by any "expansion factor" — the weights already
  perform the extrapolation.

## Building a Season-Long Dataset

```{r season-data, message = FALSE}
library(tidycreel)

set.seed(2024)

# Full season: May 1 – September 30 (184 days)
all_dates <- seq(as.Date("2024-05-01"), as.Date("2024-09-30"), by = "1 day")
day_types <- ifelse(
  as.integer(format(all_dates, "%u")) %in% c(6L, 7L), "weekend", "weekday"
)

season_calendar <- data.frame(
  date       = all_dates,
  day_type   = day_types,
  open_hours = 10
)

# Stratified random sample: 30% of weekdays, 50% of weekends
set.seed(2024)
sampled <- ave(
  seq_along(all_dates), day_types,
  FUN = function(i) {
    rate <- if (day_types[i[1]] == "weekend") 0.50 else 0.30
    as.integer(runif(length(i)) < rate)
  }
)
sampled_calendar <- season_calendar[as.logical(sampled), ]

# Generate plausible effort counts (angler-hours per count)
# Weekend counts higher than weekday
sampled_counts <- data.frame(
  date = sampled_calendar$date,
  day_type = sampled_calendar$day_type,
  n_anglers = ifelse(
    sampled_calendar$day_type == "weekend",
    round(rnorm(sum(sampled), mean = 55, sd = 18)),
    round(rnorm(sum(sampled), mean = 28, sd = 12))
  )
)
sampled_counts$n_anglers <- pmax(sampled_counts$n_anglers, 0L)

nrow(sampled_calendar) # Number of sampled days
```

## Season-Total Effort

Pass the sampled calendar to `creel_design()`, attach the counts, and call
`estimate_effort()`. The result is the estimated total angler-hours for the
entire May–September season.

```{r season-effort, message = FALSE}
season_design <- creel_design(
  sampled_calendar,
  date = date, strata = day_type
)
season_design <- add_counts(season_design, sampled_counts)

season_effort <- estimate_effort(season_design)
season_effort$estimates
```

The `estimate` column is the season-total angler-hours; `se` is the standard error of
that total.

## Monthly Effort Totals

To break the season into monthly totals, subset the calendar and counts to each
month, build a separate monthly design, and estimate. Monthly designs are independent,
so their variances can be combined later.

```{r monthly-effort, message = FALSE}
months <- 5:9
month_labels <- c("May", "June", "July", "August", "September")

monthly_effort <- lapply(seq_along(months), function(i) {
  m <- months[i]
  cal_m <- sampled_calendar[format(sampled_calendar$date, "%m") == sprintf("%02d", m), ]
  cnt_m <- sampled_counts[format(sampled_counts$date, "%m") == sprintf("%02d", m), ]

  if (nrow(cal_m) == 0) {
    return(NULL)
  }

  des_m <- creel_design(cal_m, date = date, strata = day_type)
  des_m <- add_counts(des_m, cnt_m)
  est <- estimate_effort(des_m)$estimates

  cbind(month = month_labels[i], est)
})

do.call(rbind, monthly_effort)
```

## Season Total from Monthly Estimates

Because monthly estimates are independent (non-overlapping time windows), the
season total is the sum of monthly estimates and the season variance is the sum
of monthly variances.

```{r season-from-months, message = FALSE}
monthly_df <- do.call(rbind, monthly_effort)

season_from_months <- data.frame(
  stratum  = "Season total",
  estimate = sum(monthly_df$estimate),
  se       = sqrt(sum(monthly_df$se^2))
)

season_from_months
```

This should be close to the single-design season total above (small differences are
due to independent stratification within each monthly design).

## Monthly Catch Totals

Total catch estimation follows the same monthly pattern: estimate the catch rate from
interviews, then multiply by monthly effort. First, generate synthetic interview data:

```{r interview-data, message = FALSE}
set.seed(2024)

# One interview per sampled day (simplified)
interviews <- data.frame(
  date         = sampled_calendar$date,
  day_type     = sampled_calendar$day_type,
  trip_status  = "complete",
  hours_fished = round(rnorm(nrow(sampled_calendar), mean = 3.5, sd = 1.2), 1),
  catch_total  = rpois(nrow(sampled_calendar), lambda = 1.8),
  catch_kept   = rpois(nrow(sampled_calendar), lambda = 0.6)
)
interviews$hours_fished <- pmax(interviews$hours_fished, 0.5)
```

Then loop over months exactly as for effort:

```{r monthly-catch, message = FALSE}
monthly_catch <- lapply(seq_along(months), function(i) {
  m <- months[i]
  cal_m <- sampled_calendar[format(sampled_calendar$date, "%m") == sprintf("%02d", m), ]
  cnt_m <- sampled_counts[format(sampled_counts$date, "%m") == sprintf("%02d", m), ]
  int_m <- interviews[format(interviews$date, "%m") == sprintf("%02d", m), ]

  if (nrow(cal_m) == 0) {
    return(NULL)
  }

  des_m <- creel_design(cal_m, date = date, strata = day_type)
  des_m <- add_counts(des_m, cnt_m)
  des_m <- add_interviews(des_m, int_m,
    trip_status = trip_status,
    catch = catch_total,
    effort = hours_fished,
    harvest = catch_kept
  )

  effort_est <- estimate_effort(des_m)
  catch_est <- estimate_total_catch(des_m, effort = effort_est)

  cbind(month = month_labels[i], catch_est$estimates)
})

do.call(rbind, monthly_catch)
```

## Season-Total Catch

Sum the monthly totals:

```{r season-catch}
catch_df <- do.call(rbind, monthly_catch)
season_catch <- data.frame(
  stratum  = "Season total",
  estimate = sum(catch_df$estimate),
  se       = sqrt(sum(catch_df$se^2))
)
season_catch
```

## Assembling a Summary Report

`season_summary()` assembles multiple `creel_estimates` objects into a single wide
tibble for reporting. For a monthly summary, build a list with named entries and pass
it to `season_summary()`.

```{r season-summary, message = FALSE}
# Collect effort and catch estimates for each month into named lists
effort_list <- list()
catch_list <- list()

for (i in seq_along(months)) {
  m <- months[i]
  label <- month_labels[i]

  cal_m <- sampled_calendar[format(sampled_calendar$date, "%m") == sprintf("%02d", m), ]
  cnt_m <- sampled_counts[format(sampled_counts$date, "%m") == sprintf("%02d", m), ]
  int_m <- interviews[format(interviews$date, "%m") == sprintf("%02d", m), ]

  if (nrow(cal_m) == 0) next

  des_m <- creel_design(cal_m, date = date, strata = day_type)
  des_m <- add_counts(des_m, cnt_m)
  des_m <- add_interviews(des_m, int_m,
    trip_status = trip_status,
    catch = catch_total,
    effort = hours_fished,
    harvest = catch_kept
  )

  effort_list[[label]] <- estimate_effort(des_m)
  catch_list[[label]] <- estimate_total_catch(des_m, effort = effort_list[[label]])
}

# Assemble effort report
effort_summary <- season_summary(effort_list)
effort_summary$table
```

```{r catch-summary, message = FALSE}
catch_summary <- season_summary(catch_list)
catch_summary$table
```

## Annual Totals Across Multiple Seasons

When survey data span multiple calendar years, apply the same pattern at the year level:
build one design per year, run estimators, then sum totals and combine variances.

```{r multi-year, eval = FALSE}
# Conceptual pattern — replace with actual data per year
year_effort <- list()

for (yr in c(2022, 2023, 2024)) {
  cal_yr <- subset(full_calendar, format(date, "%Y") == yr)
  cnt_yr <- subset(full_counts, format(date, "%Y") == yr)

  des_yr <- creel_design(cal_yr, date = date, strata = day_type)
  des_yr <- add_counts(des_yr, cnt_yr)

  year_effort[[as.character(yr)]] <- estimate_effort(des_yr)
}

# Multi-year report table
season_summary(year_effort)$table
```

## Exporting Results

`write_schedule()` accepts any data frame, so you can export the assembled summary
table directly:

```{r export, eval = FALSE}
write_schedule(effort_summary$table, "effort_by_month_2024.csv")
write_schedule(effort_summary$table, "effort_by_month_2024.xlsx")
```

## Key Principles

| Goal | Approach |
|------|----------|
| Season total | Single design covering full season; call `estimate_effort()` once |
| Monthly totals | Separate monthly designs; loop over months |
| Season total from months | Sum monthly estimates; sum monthly variances |
| Annual comparison | Separate annual designs; use `season_summary()` |
| Report table | `season_summary(named_list_of_estimates)$table` |

## References

- Cochran, W. G. (1977). *Sampling Techniques*, 3rd ed. Wiley, New York.

- Pollock, K. H., Jones, C. M., and Brown, T. L. (1994). *Angler Survey Methods and
  Their Applications in Fisheries Management*. American Fisheries Society, Bethesda, MD.

- Su, Y.-S., and Liu, P. (2025). Flexible creel survey estimators.
  *Canadian Journal of Fisheries and Aquatic Sciences*, 82, 1–27.
