---
title: "Getting Started with tidycreel"
output:
  rmarkdown::html_vignette:
    highlight: null
vignette: >
  %\VignetteIndexEntry{Getting Started with tidycreel}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Introduction

The tidycreel package provides a tidy, pipe-friendly interface for creel survey design and analysis. It allows fisheries biologists to work in domain vocabulary (dates, strata, counts, effort) without needing to understand the internals of the survey package. The workflow follows a simple three-step pattern:

1. **Design**: Define your survey calendar and stratification
2. **Data**: Attach count observations to the design
3. **Estimation**: Compute effort estimates with variance

## Survey Design

We start by loading the package and the example calendar dataset:

```{r design}
library(tidycreel)

# Load example calendar data
data(example_calendar)
head(example_calendar)

# Create a creel design with weekday/weekend strata
design <- creel_design(example_calendar, date = date, strata = day_type)
print(design)
```

The `creel_design()` function uses tidy selectors, so you can specify columns by name without quotes. The design object captures the survey structure: 14 days with weekday/weekend stratification.

## Adding Count Data

Next, we attach instantaneous count observations to the design:

```{r counts}
# Load example count data
data(example_counts)
head(example_counts)

# Attach counts to the design
design <- add_counts(design, example_counts)
print(design)
```

The `add_counts()` function validates that the count data matches the design structure, then constructs the internal survey design object. Notice that the design now shows count data attached with 14 observations.

## Estimating Total Effort

With count data attached, we can estimate total effort across the entire survey period:

```{r total}
# Estimate total effort
result <- estimate_effort(design)
print(result)
```

The result shows the estimated total effort, standard error, and 95% confidence interval. The estimate is approximately 358 angler-hours over the 14-day period.

## Grouped Estimation

We can also compute estimates separately for each stratum or group using the `by` parameter:

```{r grouped}
# Estimate effort by day_type
result_by_day <- estimate_effort(design, by = day_type)
print(result_by_day)
```

The grouped results show separate estimates for weekday and weekend periods. Notice that weekend effort (approximately 250 hours) is much higher than weekday effort (approximately 108 hours), reflecting typical recreational fishing patterns.

The `by` parameter accepts tidy selectors, so you can group by multiple columns or use tidyselect helpers like `starts_with()`.

## Variance Methods

By default, `estimate_effort()` uses Taylor linearization for variance estimation. The package also supports bootstrap and jackknife methods:

```{r variance}
# Bootstrap variance estimation (500 replicates)
set.seed(123) # For reproducibility
result_boot <- estimate_effort(design, variance = "bootstrap")
print(result_boot)

# Jackknife variance estimation
result_jk <- estimate_effort(design, variance = "jackknife")
print(result_jk)
```

**When to use each method:**

- **Taylor linearization** (default): Computationally efficient and appropriate for most smooth statistics. This is the recommended default.
- **Bootstrap**: Use when working with non-smooth statistics or when you want to verify Taylor linearization assumptions. More computationally intensive.
- **Jackknife**: Alternative resampling method that is deterministic (unlike bootstrap). Useful for verification or when bootstrap is too slow.

All three methods work with grouped estimation as well:

```{r grouped_variance}
# Grouped estimation with bootstrap variance
set.seed(123)
result_grouped_boot <- estimate_effort(design, by = day_type, variance = "bootstrap")
print(result_grouped_boot)
```

## Schedule-Defined Special Strata

If your survey calendar includes prospective high-use or other special periods, the same
three-step workflow still applies. The difference is that the schedule may carry a resolved
`final_stratum` column from `generate_schedule(..., special_periods = ...)`, and that
resolved stratum should drive the analysis design.

```{r special-strata-overview, eval = FALSE}
sched <- generate_schedule(
  start_date = "2027-07-24",
  end_date = "2027-08-04",
  n_periods = 1,
  sampling_rate = 0.5,
  include_all = TRUE,
  special_periods = opener_periods,
  seed = 42
)

calendar_for_design <- transform(
  sched[, c("date", "final_stratum")],
  analysis_stratum = ifelse(grepl("^high_use", final_stratum), final_stratum, "regular")
)[, c("date", "analysis_stratum")]

design_special <- creel_design(
  calendar_for_design,
  date = date,
  strata = analysis_stratum
)
```

Once counts and interviews are attached, `estimate_effort(..., target = "period_total")`
and the total-catch/product estimators use the declared analysis strata directly. If one of
those strata is too sparse for variance estimation, tidycreel names the sparse stratum in
its diagnostic instead of failing opaquely.

## Next Steps

This vignette covers the core tidycreel workflow for instantaneous count surveys. For more details on specific functions, see their help pages:

- `?creel_design` - Define survey calendar and stratification
- `?add_counts` - Attach count data to a design
- `?estimate_effort` - Compute effort estimates with variance
- `?as_survey_design` - Extract internal survey object for advanced use

For information on the example datasets:

- `?example_calendar` - Example survey calendar
- `?example_counts` - Example count observations
