---
title: "Unextrapolated Interview Summaries"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Unextrapolated Interview Summaries}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Introduction

Creel surveys produce two types of summary products: **unextrapolated summaries** and
**extrapolated estimates**. This vignette covers unextrapolated summaries — raw tabulations
of interview records that describe the composition of interviewed parties without applying
survey-design weighting.

**When to use unextrapolated summaries:**

- Describing who was interviewed (angler type, method, species sought)
- Checking for refusals and participation rates
- Examining trip-length distributions
- Computing catch-while-sought and harvest-while-sought rates as quality indicators

**Important caveat:** All unextrapolated functions are *interview-weighted*, not
*pressure-weighted*. A day with many anglers and few interviews has the same weight
as a day with few anglers and many interviews. For pressure-weighted, design-correct
estimates, use `estimate_cpue()`, `estimate_total_catch()`, and related functions
described in the "Interview-Based Catch Estimation" vignette.

## Build the Design

Start by assembling a complete design using all v0.5.0 data layers:

```{r build-design, message = FALSE}
library(tidycreel)

# Load example datasets
data(example_calendar)
data(example_counts)
data(example_interviews)
data(example_catch)
data(example_lengths)

# Step 1: Define the survey calendar and stratification
design <- creel_design(example_calendar, date = date, strata = day_type)

# Step 2: Attach instantaneous count observations
design <- add_counts(design, example_counts)

# Step 3: Attach interview records (all extended v0.5.0 fields)
design <- add_interviews(design, example_interviews,
  catch          = catch_total,
  effort         = hours_fished,
  harvest        = catch_kept,
  trip_status    = trip_status,
  trip_duration  = trip_duration,
  angler_type    = angler_type,
  angler_method  = angler_method,
  species_sought = species_sought,
  n_anglers      = n_anglers,
  refused        = refused
)

# Step 4: Attach species-level catch data
design <- add_catch(design, example_catch,
  catch_uid     = interview_id,
  interview_uid = interview_id,
  species       = species,
  count         = count,
  catch_type    = catch_type
)

# Step 5: Attach fish length data (individual harvest + binned release)
design <- add_lengths(design, example_lengths,
  length_uid     = interview_id,
  interview_uid  = interview_id,
  species        = species,
  length         = length,
  length_type    = length_type,
  count          = count,
  release_format = "binned"
)

print(design)
```

The `print()` output shows all five data layers: calendar, counts, interviews, catch, and lengths.

## Interview Participation

Use `summarize_refusals()` to understand how many potential interviewees declined to
participate. High refusal rates can bias estimates if refusers differ systematically
from participants.

```{r refusals}
summarize_refusals(design)
```

All 22 interviews in `example_interviews` were accepted (no refusals). In field surveys,
a refusal rate above 20% warrants investigation.

## Interview Composition

Seven tabulation functions describe the composition of the interview sample.

### Day Type

```{r day-type}
summarize_by_day_type(design)
```

### Angler Type

```{r angler-type}
summarize_by_angler_type(design)
```

### Fishing Method

```{r method}
summarize_by_method(design)
```

### Species Sought

```{r species-sought}
summarize_by_species_sought(design)
```

### Successful Parties

`summarize_successful_parties()` counts parties that caught at least one fish of their
target species (as recorded in catch data), broken down by angler type and species sought.

```{r successful}
summarize_successful_parties(design)
```

### Trip Length Distribution

```{r trip-length}
summarize_by_trip_length(design)
```

The trip-length distribution is useful for detecting outliers and understanding effort
patterns.

## Catch-While-Sought (CWS) and Harvest-While-Sought (HWS) Rates

CWS and HWS rates measure the proportion of interviews where anglers targeting a species
actually caught (CWS) or kept (HWS) that species. These are useful quality indicators and
input parameters for population models, but should **not** be treated as pressure-weighted
catch rates for management use — they reflect the interview sample composition, not fishing
pressure across the survey period.

### CWS Rates by Species Sought

```{r cws}
summarize_cws_rates(design, by = species_sought)
```

### CWS Rates Collapsed Across All Groupings

```{r cws-collapsed}
summarize_cws_rates(design, by = NULL)
```

### HWS Rates

```{r hws}
summarize_hws_rates(design, by = species_sought)
```

**Interpretation guidance:** CWS and HWS rates > 1.0 are possible when anglers catch
multiple fish of the target species in a single interview. A CWS rate of 0.6 means
60% of interviews targeting that species resulted in at least one catch.

## Length Frequency Distributions

Length data attached via `add_lengths()` can be summarized by catch type and species.

### Harvest Lengths

```{r lfreq-harvest}
summarize_length_freq(design, type = "harvest", by = species, bin_width = 25)
```

### Release Lengths

Release lengths in `example_lengths` are stored in pre-binned format.
`summarize_length_freq()` handles this automatically:

```{r lfreq-release}
summarize_length_freq(design, type = "release", by = species)
```

### All-Catch Lengths

```{r lfreq-catch}
summarize_length_freq(design, type = "catch", by = species, bin_width = 25)
```

## Extrapolated Species-Level Estimates

For design-correct, pressure-weighted estimates broken down by species, use the
extrapolated estimators. These combine effort estimates (from count data) with
species-specific catch rates (from interview + catch data) using the ratio-of-means
estimator.

### Catch Per Unit Effort by Species

```{r cpue-species, message = FALSE}
estimate_cpue(design, by = species)
```

### Total Catch by Species

```{r total-catch-species, message = FALSE}
estimate_total_catch(design, by = species)
```

### Total Harvest by Species

```{r total-harvest-species, message = FALSE}
estimate_total_harvest(design, by = species)
```

### Release Rate and Total Releases by Species

```{r release-species, message = FALSE}
estimate_release_rate(design, by = species)
estimate_total_release(design, by = species)
```

For grouped estimates combining calendar strata with species, use
`by = c(day_type, species)`. See `vignette("interview-estimation")` for the complete
extrapolated estimation workflow.

## Summary

| Function | Data required | Output |
|----------|---------------|--------|
| `summarize_refusals()` | refused field | month × participation × N × % |
| `summarize_by_day_type()` | strata | month × day_type × N × % |
| `summarize_by_angler_type()` | angler_type | month × angler_type × N × % |
| `summarize_by_method()` | angler_method | month × method × N × % |
| `summarize_by_species_sought()` | species_sought | month × species × N × % |
| `summarize_successful_parties()` | catch + angler_type + species_sought | angler_type × species × success rate |
| `summarize_by_trip_length()` | trip_duration | bin × N × % |
| `summarize_cws_rates()` | catch + species_sought | CWS rate ± SE by grouping |
| `summarize_hws_rates()` | catch + species_sought | HWS rate ± SE by grouping |
| `summarize_length_freq()` | lengths | bin × N × % × cumulative % |
