---
title: "Visualising Creel Survey Designs and Results"
output:
  rmarkdown::html_vignette:
    highlight: null
vignette: >
  %\VignetteIndexEntry{Visualising Creel Survey Designs and Results}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4,
  out.width = "100%"
)
```

tidycreel ships three ggplot2-based plotting functions that cover the main
inspection points in a creel survey workflow:

| Function | When to use |
|----------|-------------|
| `plot_design()` | Inspect stratum sample sizes and count distributions |
| `autoplot(schedule)` | Review the survey calendar tile-by-tile |
| `autoplot(estimates)` | Visualise effort or CPUE estimates with CIs |

```{r load}
library(tidycreel)
library(ggplot2)
```

---

## 1  Inspect the design with `plot_design()`

### Before attaching counts

Once you have built a `creel_design` object from a calendar, `plot_design()`
shows the number of sampled days per stratum.

```{r design-no-counts}
# Build a design from the bundled example calendar
data("example_counts")
cal <- unique(example_counts[, c("date", "day_type")])
design <- creel_design(cal, date = date, strata = day_type)

plot_design(design, title = "Sample sizes by stratum")
```

The bar chart makes it immediately clear whether weekday and weekend strata
are balanced. A tall weekday bar with a short weekend bar signals that
estimates for the weekend stratum will carry higher uncertainty.

### After attaching counts

Once counts are attached, `plot_design()` switches to a jitter + crossbar
display showing the raw count distribution per stratum.

```{r design-with-counts, warning = FALSE}
design <- add_counts(design, example_counts)

plot_design(design, title = "Count distribution by stratum")
```

The crossbar shows the mean with a 95 % normal CI across sampled days.
Large within-stratum spread (many jittered points far from the crossbar)
suggests high day-to-day variability and the need for more sampling days.

---

## 2  Review the survey calendar with `autoplot()`

`autoplot.creel_schedule()` renders a monthly tile calendar from a
`creel_schedule` object — the same object produced by `generate_schedule()`.

```{r schedule-plot}
# Generate a three-month schedule sampling 40 % of days
schedule <- generate_schedule(
  start_date    = "2024-06-01",
  end_date      = "2024-08-31",
  n_periods     = 2,
  sampling_rate = 0.4,
  seed          = 42
)

autoplot(schedule, title = "2024 Summer Creel Schedule")
```

Blue tiles are sampled **weekdays**; red tiles are sampled **weekends**;
grey tiles are unsampled days. Month panels stack vertically so the full
season is visible at a glance.

---

## 3  Visualise estimates with `autoplot()`

`autoplot.creel_estimates()` draws a point-and-errorbar plot from any
`creel_estimates` object.

### Ungrouped effort estimate

```{r effort-ungrouped, warning = FALSE}
data("example_interviews")

design <- add_interviews(
  design,
  example_interviews,
  catch       = catch_total,
  effort      = hours_fished,
  trip_status = trip_status
)

effort <- estimate_effort(design)

autoplot(effort, title = "Total angler-effort estimate")
```

### Grouped effort estimate

Passing `by = day_type` estimates effort separately for each stratum. The
resulting plot maps the grouping variable to both the x-axis and point
colour.

```{r effort-grouped, warning = FALSE}
effort_by_type <- estimate_effort(design, by = day_type)

autoplot(effort_by_type, title = "Angler-effort by day type")
```

Weekend effort is substantially higher than weekday effort — a typical
pattern in summer recreational fisheries.

### CPUE estimate

```{r cpue, warning = FALSE}
cpue <- estimate_catch_rate(design)

autoplot(cpue, title = "Walleye CPUE (catch per angler-hour)")
```

---

## Combining plots

All three functions return standard `ggplot` objects, so they compose
naturally with `+` (ggplot2 operators) or side-by-side using
[patchwork](https://patchwork.data-imaginist.com/) if that package is
installed.

```{r combine, eval = FALSE}
# Requires patchwork
library(patchwork)
plot_design(design) + autoplot(effort)
```

---

## Summary

| Function | Input class | Returns |
|----------|-------------|---------|
| `plot_design(design)` | `creel_design` | bar chart (no counts) or jitter+crossbar (counts) |
| `autoplot(schedule)` | `creel_schedule` | monthly tile calendar |
| `autoplot(estimates)` | `creel_estimates` | point-and-errorbar plot |

All plots accept a `title =` argument and return a `ggplot` object for
further customisation with standard ggplot2 `+` syntax.
