---
title: "Creel Report"
subtitle: "A tidycreel Quarto starter template"
author: "Your Name"
date: today
format:
  pdf:
    toc: true
    toc-depth: 3
    number-sections: true
    colorlinks: true
    geometry: margin=1in
  html:
    toc: true
    toc-depth: 3
    number-sections: true
    theme: [cosmo, theme.scss]
execute:
  echo: false
  warning: false
  message: false
  freeze: auto
---

```{r setup}
library(tidycreel)
library(ggplot2)
library(knitr)

data(example_calendar)
data(example_counts)
data(example_interviews)
data(example_catch)
data(example_lengths)

# Replace these example objects with your own project data imports.
calendar <- example_calendar
counts <- example_counts
interviews <- example_interviews
catch <- example_catch
lengths <- example_lengths

design <- creel_design(calendar, date = date, strata = day_type)
design <- add_counts(design, counts)
design <- add_interviews(
  design,
  interviews,
  catch = catch_total,
  effort = hours_fished,
  harvest = catch_kept,
  trip_status = trip_status,
  trip_duration = trip_duration,
  angler_type = angler_type,
  angler_method = angler_method,
  species_sought = species_sought,
  n_anglers = n_anglers,
  refused = refused
)
design <- add_catch(
  design,
  catch,
  catch_uid = interview_id,
  interview_uid = interview_id,
  species = species,
  count = count,
  catch_type = catch_type
)
design <- add_lengths(
  design,
  lengths,
  length_uid = interview_id,
  interview_uid = interview_id,
  species = species,
  length = length,
  length_type = length_type,
  count = count,
  release_format = "binned"
)

design_table <- data.frame(
  Metric = c(
    "Survey type",
    "Date column",
    "Strata",
    "Calendar days",
    "Calendar range",
    "Count records",
    "Interview records",
    "Catch records",
    "Length records"
  ),
  Value = c(
    design$design_type,
    design$date_col,
    paste(design$strata_cols, collapse = ", "),
    nrow(design$calendar),
    paste(range(design$calendar[[design$date_col]]), collapse = " to "),
    nrow(design$counts),
    nrow(design$interviews),
    nrow(design$catch),
    nrow(design$lengths)
  ),
  stringsAsFactors = FALSE
)

maybe <- function(expr, required = character()) {
  if (length(required) > 0L && !all(required %in% names(design$interviews))) {
    return(NULL)
  }
  tryCatch(expr, error = function(e) NULL)
}

validation <- validation_report(
  counts = counts,
  interviews = interviews
)
trip_summary <- summarize_trips(design)
refusal_summary <- maybe(summarize_refusals(design))
day_type_summary <- maybe(summarize_by_day_type(design), required = "day_type")
angler_type_summary <- maybe(summarize_by_angler_type(design), required = "angler_type")
method_summary <- maybe(summarize_by_method(design), required = "angler_method")
species_sought_summary <- maybe(
  summarize_by_species_sought(design),
  required = "species_sought"
)
trip_length_summary <- maybe(summarize_by_trip_length(design), required = "trip_duration")
successful_parties_summary <- maybe(
  summarize_successful_parties(design),
  required = c("angler_type", "species_sought")
)
cws_month_species_summary <- maybe(
  summarize_cws_rates(design, by = c(month, species_sought)),
  required = "species_sought"
)
cws_group_summary <- maybe(
  summarize_cws_rates(design, by = c(month, angler_type, species_sought)),
  required = c("angler_type", "species_sought")
)
hws_month_species_summary <- maybe(
  summarize_hws_rates(design, by = c(month, species_sought)),
  required = "species_sought"
)
hws_group_summary <- maybe(
  summarize_hws_rates(design, by = c(month, angler_type, species_sought)),
  required = c("angler_type", "species_sought")
)
length_catch_summary <- maybe(
  summarize_length_freq(design, type = "catch", by = species, bin_width = 25),
  required = "species"
)
length_harvest_summary <- maybe(
  summarize_length_freq(design, type = "harvest", by = species, bin_width = 25),
  required = "species"
)
length_release_summary <- maybe(
  summarize_length_freq(design, type = "release", by = species, bin_width = 25),
  required = "species"
)

effort_total <- estimate_effort(design)
effort_by_day <- estimate_effort(design, by = day_type)
angler_trips <- estimate_angler_trips(effort_total, design)
catch_rate_by_species <- maybe(
  estimate_catch_rate(design, by = species_sought),
  required = "species_sought"
)
harvest_rate_by_species <- maybe(
  estimate_harvest_rate(design, by = species_sought),
  required = "species_sought"
)
release_rate_by_species <- maybe(
  estimate_release_rate(design, by = species_sought),
  required = "species_sought"
)
total_catch <- estimate_total_catch(design)
total_harvest <- estimate_total_harvest(design)
total_release <- estimate_total_release(design)
season_totals <- season_summary(list(
  effort = effort_total,
  trips = angler_trips,
  catch = total_catch,
  harvest = total_harvest,
  release = total_release
))
```

## Design Parameters

Replace the example data imports in the setup chunk with your own survey tables, then keep the report structure and `tidycreel` calls intact.

```{r design-parameters}
knitr::kable(design_table, col.names = c("Metric", "Value"))
```

## Data Validation

```{r validation-report}
knitr::kable(as.data.frame(validation))
```

## Unextrapolated Summaries

### Trip Summary

```{r trip-summary}
knitr::kable(
  data.frame(
    Metric = c("Total interviews", "Complete trips", "Incomplete trips", "Complete share"),
    Value = c(
      trip_summary$n_total,
      trip_summary$n_complete,
      trip_summary$n_incomplete,
      sprintf("%.1f%%", trip_summary$pct_complete)
    ),
    stringsAsFactors = FALSE
  )
)
```

### Refusals

```{r refusals}
if (!is.null(refusal_summary)) knitr::kable(as.data.frame(refusal_summary))
```

### By Day Type

```{r by-day-type}
if (!is.null(day_type_summary)) knitr::kable(as.data.frame(day_type_summary))
```

### By Angler Type

```{r by-angler-type}
if (!is.null(angler_type_summary)) knitr::kable(as.data.frame(angler_type_summary))
```

### By Angler Method

```{r by-method}
if (!is.null(method_summary)) knitr::kable(as.data.frame(method_summary))
```

### By Species Sought

```{r by-species-sought}
if (!is.null(species_sought_summary)) knitr::kable(as.data.frame(species_sought_summary))
```

### Successful Parties

```{r successful-parties}
if (!is.null(successful_parties_summary)) knitr::kable(as.data.frame(successful_parties_summary))
```

### Trip Length

```{r trip-length}
if (!is.null(trip_length_summary)) knitr::kable(as.data.frame(trip_length_summary))
```

## Length Frequencies

### Catch

```{r length-catch}
if (!is.null(length_catch_summary)) knitr::kable(as.data.frame(length_catch_summary))
```

### Harvest

```{r length-harvest}
if (!is.null(length_harvest_summary)) knitr::kable(as.data.frame(length_harvest_summary))
```

### Release

```{r length-release}
if (!is.null(length_release_summary)) knitr::kable(as.data.frame(length_release_summary))
```

## Caught-while-sought and Harvest-while-sought

### CWS Rates by Month and Species Sought

```{r cws-month-species}
if (!is.null(cws_month_species_summary)) knitr::kable(as.data.frame(cws_month_species_summary))
```

### CWS Rates by Month, Angler Type, and Species Sought

```{r cws-group}
if (!is.null(cws_group_summary)) knitr::kable(as.data.frame(cws_group_summary))
```

### HWS Rates by Month and Species Sought

```{r hws-month-species}
if (!is.null(hws_month_species_summary)) knitr::kable(as.data.frame(hws_month_species_summary))
```

### HWS Rates by Month, Angler Type, and Species Sought

```{r hws-group}
if (!is.null(hws_group_summary)) knitr::kable(as.data.frame(hws_group_summary))
```

## Extrapolated Estimates

### Effort

```{r effort-total}
ggplot2::autoplot(
  effort_total,
  title = "Estimated total effort",
  theme = "creel"
)
```

```{r effort-by-day}
ggplot2::autoplot(
  effort_by_day,
  title = "Effort by day type",
  theme = "creel"
)
```

### Angler Trips

```{r angler-trips}
ggplot2::autoplot(
  angler_trips,
  title = "Estimated angler trips",
  theme = "creel"
)
```

### Catch, Harvest, and Release

```{r catch-rate}
if (!is.null(catch_rate_by_species)) {
  ggplot2::autoplot(
    catch_rate_by_species,
    title = "Catch rate by species sought",
    theme = "creel"
  )
}
```

```{r harvest-rate}
if (!is.null(harvest_rate_by_species)) {
  ggplot2::autoplot(
    harvest_rate_by_species,
    title = "Harvest rate by species sought",
    theme = "creel"
  )
}
```

```{r release-rate}
if (!is.null(release_rate_by_species)) {
  ggplot2::autoplot(
    release_rate_by_species,
    title = "Release rate by species sought",
    theme = "creel"
  )
}
```

```{r total-estimates}
knitr::kable(season_totals$table)
```

## Notes

- Use the example data as a working starting point, then swap in your own calendar, count, interview, catch, and length tables.
- Keep the `creel_design()` through `season_summary()` workflow so the report stays aligned with `tidycreel`.
- Prefer `theme = "creel"` for plots so the report matches package styling.
