---
title: "Getting Started with tidycreel.connect"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with tidycreel.connect}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>"
)
```

## Overview

**tidycreel.connect** provides a unified connection layer for the
[tidycreel](https://github.com/cchizinski2/tidycreel) creel survey package.
It lets you load interview, count, catch, and length data from two backends:

| Backend | When to use |
|---------|-------------|
| **CSV** | Flat files from a data export or field laptop |
| **SQL Server** | Agency production database (e.g. NGPC creel DB) |

The same `creel_connection` object is accepted by all `fetch_*()` functions
and passes cleanly into `tidycreel::add_interviews()`.

## Installation prerequisites

```{r prereqs, eval = FALSE}
# Core package (already installed if you have tidycreel)
install.packages("tidycreel.connect")

# Required for YAML-based connections
install.packages("config")

# Required for SQL Server connections
install.packages("odbc")
```

You also need a native ODBC driver on your OS — see the README for
platform-specific install instructions.

---

## 1. Define a column-mapping schema

A `creel_schema` tells tidycreel.connect which columns in your data files
correspond to the canonical names used by `tidycreel`. Create one with
`tidycreel::creel_schema()`:

```{r schema, eval = TRUE}
library(tidycreel)
library(tidycreel.connect)

schema <- creel_schema(
  survey_type       = "instantaneous",
  # Table names (used by SQL Server backend; ignored for CSV backend)
  interviews_table  = "vwInterviews",
  counts_table      = "vwCounts",
  catch_table       = "vwCatch",
  lengths_table     = "vwLengths",
  # Column mappings: right-hand side is the name in your data source
  interview_uid_col = "InterviewID",
  date_col          = "SurveyDate",
  effort_col        = "EffortHours",
  catch_col         = "TotalCatch",
  trip_status_col   = "TripStatus",
  count_col         = "AnglerCount",
  catch_uid_col     = "CatchUID",
  species_col       = "SpeciesCode",
  catch_count_col   = "CatchCount",
  catch_type_col    = "CatchType",
  length_uid_col    = "LengthUID",
  length_mm_col     = "LengthMM",
  length_type_col   = "LengthType"
)

schema
```

---

## 2a. Connect using a YAML config file (recommended)

YAML config keeps connection details out of your scripts. Create a file
`creel-config.yml`:

```yaml
# creel-config.yml — CSV backend example
default:
  backend: csv
  files:
    interviews:      data/interviews.csv
    counts:          data/counts.csv
    catch:           data/catch.csv
    harvest_lengths: data/harvest_lengths.csv
    release_lengths: data/release_lengths.csv
  schema:
    survey_type: instantaneous
```

Then connect:

```{r yaml-csv, eval = FALSE}
conn <- creel_connect_from_yaml("creel-config.yml")
```

### SQL Server YAML config

Credentials must **never** be stored as plain text. Use the `config` package's
`!expr` tag to inject them from environment variables:

```yaml
# creel-config.yml — SQL Server backend example
default:
  backend:   sqlserver
  server:    myserver.example.com
  database:  CreelDB
  auth_type: password
  username:  !expr Sys.getenv("CREEL_USER")
  password:  !expr Sys.getenv("CREEL_PASS")
  schema:
    survey_type:       instantaneous
    interviews_table:  vwInterviews
    counts_table:      vwCounts
    catch_table:       vwCatch
    lengths_table:     vwLengths
    interview_uid_col: InterviewID
    date_col:          SurveyDate
    effort_col:        EffortHours
    catch_col:         TotalCatch
    trip_status_col:   TripStatus
    count_col:         AnglerCount
    catch_uid_col:     CatchUID
    species_col:       SpeciesCode
    catch_count_col:   CatchCount
    catch_type_col:    CatchType
    length_uid_col:    LengthUID
    length_mm_col:     LengthMM
    length_type_col:   LengthType
```

Set credentials before connecting:

```{r yaml-sqlserver, eval = FALSE}
Sys.setenv(CREEL_USER = "your_username", CREEL_PASS = "your_password")
conn <- creel_connect_from_yaml("creel-config.yml")
```

For Windows integrated authentication (Windows only):

```yaml
auth_type: windows
```

> **Note:** `auth_type: windows` aborts immediately on macOS and Linux with
> a helpful error message directing you to use `auth_type: password` instead.

### Multiple environments

The `config` environment block lets you switch between development and
production databases without editing the file:

```yaml
default:
  backend: csv
  files:
    interviews: data/dev/interviews.csv
    # ...

production:
  backend:   sqlserver
  server:    prod.example.com
  database:  CreelDB
  auth_type: password
  username:  !expr Sys.getenv("CREEL_USER")
  password:  !expr Sys.getenv("CREEL_PASS")
  # ...
```

```{r multi-env, eval = FALSE}
# Default (CSV for development)
conn_dev <- creel_connect_from_yaml("creel-config.yml")

# Production SQL Server
conn_prod <- creel_connect_from_yaml("creel-config.yml", config = "production")
```

---

## 2b. Connect programmatically

You can also construct a connection directly without a YAML file.

### CSV backend

```{r connect-csv, eval = TRUE}
# Build small toy CSV files for this example
tmp_dir <- withr::local_tempdir()

interviews_csv <- file.path(tmp_dir, "interviews.csv")
counts_csv <- file.path(tmp_dir, "counts.csv")
catch_csv <- file.path(tmp_dir, "catch.csv")
lengths_csv <- file.path(tmp_dir, "lengths.csv")

readr::write_csv(
  data.frame(
    InterviewID  = 1:4,
    SurveyDate   = "2024-06-01",
    EffortHours  = c(2.0, 3.5, 1.5, 4.0),
    TotalCatch   = c(5L, 8L, 2L, 6L),
    TripStatus   = "complete"
  ),
  interviews_csv
)
readr::write_csv(
  data.frame(SurveyDate = "2024-06-01", AnglerCount = 12L),
  counts_csv
)
readr::write_csv(
  data.frame(
    CatchUID    = 1:3,
    InterviewID = c(1L, 1L, 2L),
    SpeciesCode = c("WAE", "WAE", "LMB"),
    CatchCount  = c(2L, 1L, 3L),
    CatchType   = "harvest"
  ),
  catch_csv
)
readr::write_csv(
  data.frame(
    LengthUID   = 1:2,
    InterviewID = c(1L, 2L),
    SpeciesCode = c("WAE", "LMB"),
    LengthMM    = c(420L, 380L),
    LengthType  = c("harvest", "harvest")
  ),
  lengths_csv
)

# Connect with a named list of file paths
conn <- creel_connect(
  con = list(
    interviews      = interviews_csv,
    counts          = counts_csv,
    catch           = catch_csv,
    harvest_lengths = lengths_csv,
    release_lengths = lengths_csv
  ),
  schema = schema
)

conn
```

### SQL Server / DBI backend

```{r connect-dbi, eval = FALSE}
library(DBI)
library(odbc)

dbi_con <- dbConnect(
  odbc(),
  driver   = "ODBC Driver 17 for SQL Server",
  server   = "myserver.example.com",
  database = "CreelDB",
  uid      = Sys.getenv("CREEL_USER"),
  pwd      = Sys.getenv("CREEL_PASS")
)

conn <- creel_connect(dbi_con, schema, auth_type = "password")
```

---

## 3. Fetch data

Once you have a `creel_connection`, use the `fetch_*()` family to load each
table. All functions return a tibble with canonical column names defined by
the schema.

```{r fetch, eval = TRUE}
interviews <- fetch_interviews(conn)
counts <- fetch_counts(conn)
catch_data <- fetch_catch(conn)
harvest_lengths <- fetch_harvest_lengths(conn)
release_lengths <- fetch_release_lengths(conn)

# Date columns are coerced to R Date regardless of backend
class(interviews$date)

# Species codes are coerced to character (SQL Server may store as integer)
class(catch_data$species)

interviews
```

---

## 4. Integrate with tidycreel

The fetched tibbles slot directly into the tidycreel estimation workflow.
`fetch_interviews()` returns a tibble with canonical column names — pass it
to `tidycreel::add_interviews()` along with a `creel_design` object:

```{r add-interviews, eval = FALSE}
# First, build a creel_design from your calendar and count data
# (see the tidycreel getting-started vignette for the full workflow)
design <- creel_design(
  calendar    = my_calendar,
  date        = date,
  strata      = stratum,
  design_type = "instantaneous"
)

# Then attach the fetched interview data
design <- add_interviews(
  design,
  interviews  = interviews,
  catch       = catch_count,
  effort      = effort,
  trip_status = trip_status
)

# Now estimate
estimate_effort(design)
estimate_catch_rate(design, species = "WAE")
```

The column names (`catch_count`, `effort`, `trip_status`) come from the
`creel_schema` column mappings defined earlier — they are the canonical names
that `fetch_interviews()` renames to, regardless of what your source database
calls them.

See `vignette("getting-started", package = "tidycreel")` for a complete
end-to-end estimation walkthrough.

---

## Credential security

Never store credentials as plain text in scripts or YAML files. The
recommended approach:

1. **Store in `.Renviron`** (user-level, never committed):
   ```
   CREEL_USER=your_username
   CREEL_PASS=your_password
   ```

2. **Reference via `!expr` in YAML**:
   ```yaml
   username: !expr Sys.getenv("CREEL_USER")
   password: !expr Sys.getenv("CREEL_PASS")
   ```

3. **Reload without restarting R** after editing `.Renviron`:
   ```{r renviron, eval = FALSE}
readRenviron("~/.Renviron")
   ```

> **Tip:** Add `.Renviron` to your global `.gitignore` to prevent accidental
> credential commits.
